---
title: "Stewart and Greenhalgh: Shadow Prices and Agricultural Land"
# toc: true
format:
  pdf:
    papersize: letter
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup-model
#| include: false
# Requires lpSolve and knitr. All coefficients are from Table 1.
library(lpSolve)

# Columns: C1, W1, P1, C2, W2, P2, C3, W3, P3.
cost <- c(66.40, 96.40, 80.14, 151.20, 93.80, 84.16,
          102.60, 61.50, 75.48)
names(cost) <- c("C1", "W1", "P1", "C2", "W2", "P2",
                 "C3", "W3", "P3")
A <- rbind(
  corn_output    = c(80, 0, 0, 60, 0, 0, 40, 0, 0),
  wheat_output   = c(0, 40, 0, 0, 35, 0, 0, 25, 0),
  pasture_output = c(0, 0, 2.2, 0, 0, 2.0, 0, 0, 1.8),
  land1          = c(1, 1, 1, 0, 0, 0, 0, 0, 0),
  corn1          = c(1, 0, 0, 0, 0, 0, 0, 0, 0),
  wheat1         = c(0, 1, 0, 0, 0, 0, 0, 0, 0),
  pasture1       = c(0, 0, 1, 0, 0, 0, 0, 0, 0),
  land2          = c(0, 0, 0, 1, 1, 1, 0, 0, 0),
  corn2          = c(0, 0, 0, 1, 0, 0, 0, 0, 0),
  wheat2         = c(0, 0, 0, 0, 1, 0, 0, 0, 0),
  pasture2       = c(0, 0, 0, 0, 0, 1, 0, 0, 0),
  land3          = c(0, 0, 0, 0, 0, 0, 1, 1, 1),
  corn3          = c(0, 0, 0, 0, 0, 0, 1, 0, 0),
  wheat3         = c(0, 0, 0, 0, 0, 0, 0, 1, 0),
  pasture3       = c(0, 0, 0, 0, 0, 0, 0, 0, 1)
)
directions <- c(rep("=", 3), rep(c("<=", "<=", "<=", ">="), 3))
package_rows <- match(c("land2", "corn2", "wheat2", "pasture2"),
                      rownames(A))
package_weights <- c(1, 0.4, 0.3, 0.3)

solve_land <- function(L = 100) {
  rhs <- c(7500, 2500, 170, 100, 60, 20, 20,
           L, 0.4 * L, 0.3 * L, 0.3 * L, 100, 20, 30, 30)
  fit <- lp("min", cost, A, directions, rhs, compute.sens = TRUE)
  if (fit$status != 0) stop("No optimal solution at L = ", L)

  # Audit the original constraints and reconstruct the objective.
  residual <- drop(A %*% fit$solution) - rhs
  stopifnot(all(fit$solution >= -1e-7),
            all(abs(residual[directions == "="]) < 1e-6),
            all(residual[directions == "<="] < 1e-6),
            all(residual[directions == ">="] > -1e-6),
            abs(sum(cost * fit$solution) - fit$objval) < 1e-6)
  list(cost = fit$objval, acres = fit$solution,
       dual = fit$duals[seq_len(nrow(A))],
       mv = -sum(fit$duals[package_rows] * package_weights))
}

baseline <- solve_land(100)
epsilon <- 0.001
finite_difference <- (baseline$cost - solve_land(100 + epsilon)$cost) /
  epsilon
stopifnot(abs(finite_difference - baseline$mv) < 1e-5)

# Compute the curve over a feasible interval, including its negative portion.
land_grid <- seq(90.5, 120, by = 0.025)
grid_fits <- lapply(land_grid, solve_land)
grid_mv <- vapply(grid_fits, function(s) s$mv, numeric(1))

# Within each LP regime, C*(L) = intercept - MV * L.
# Locate the corners by intersecting adjacent cost segments, then verify
# that the resulting piecewise-linear function matches every grid solution.
regime_starts <- c(1L, which(abs(diff(grid_mv)) > 1e-6) + 1L)
regime_mv <- grid_mv[regime_starts]
intercepts <- vapply(grid_fits[regime_starts], function(s) s$cost,
                     numeric(1)) + regime_mv * land_grid[regime_starts]
breaks <- diff(intercepts) / diff(regime_mv)
edges <- c(min(land_grid), breaks, max(land_grid))
segment_id <- pmin(findInterval(land_grid, edges), length(regime_mv))
predicted_cost <- intercepts[segment_id] -
  regime_mv[segment_id] * land_grid
actual_cost <- vapply(grid_fits, function(s) s$cost, numeric(1))
stopifnot(all(diff(edges) > 0),
          max(abs(predicted_cost - actual_cost)) < 1e-5)

# Check dual-based slopes against local re-solves inside each regime.
for (L in (head(edges, -1) + tail(edges, -1)) / 2) {
  s <- solve_land(L)
  fd <- (s$cost - solve_land(L + epsilon)$cost) / epsilon
  stopifnot(abs(fd - s$mv) < 1e-5)
}

money <- function(x) formatC(x, format = "f", digits = 2, big.mark = ",")
```



# Motivation: What is an additional acre worth?

**Reading:** Fred J. Stewart and Richard A. Greenhalgh (1977), *The Use of
Shadow Prices in Determining Marginal Values for Agricultural Land*.
The numerical model below uses Table 1 of the paper; the resource-package
argument appears on printed pages 3--7 and in footnote 1.

Land development projects, including drainage, flood protection, and erosion
control, can expand the agricultural resource base. A planner needs to compare
the benefits of those additions with their development costs. Market evidence
can be limited, and a land price need not reveal the contribution of another
acre to a particular production system.

Stewart and Greenhalgh ask how an optimization model can estimate that marginal
contribution. Their example holds agricultural production requirements fixed
and minimizes the cost of meeting them. Additional useful land creates value
when it allows the same output to be produced at lower cost.

The complication is that **land is a package of characteristics**. An acre can
bring a total-acreage allowance, limits on suitable crop uses, and a requirement
to maintain some pasture. Expanding the resource changes several constraints
at once. The shadow price on the total-land constraint values only one of those
changes.

This application connects duality to an economic question: *Which right-hand
sides actually change when a project adds a unit of the resource?* The answer
determines which shadow prices belong in its marginal valuation.

The paper develops a small illustrative model, rather than estimating a model
from a documented farm dataset. Its values are conditional on the stated
production requirements, technologies, costs, and land-use restrictions. They
measure production-cost savings over the model's accounting period, not a
capitalized market sale price for land.

# The agricultural production model

## Activities and costs

There are three land groups, indexed by $g=1,2,3$, and three activities on each
group. Let $C_g$, $W_g$, and $P_g$ denote acres of corn, wheat, and pasture,
respectively. These nine acreages are the baseline decision variables.

The full numerical objective is

$$
\begin{aligned}
\min_{C_1,W_1,P_1,C_2,W_2,P_2,C_3,W_3,P_3}\quad
&66.40C_1+96.40W_1+80.14P_1\\
&{}+151.20C_2+93.80W_2+84.16P_2\\
&{}+102.60C_3+61.50W_3+75.48P_3.
\end{aligned}
$$

Each coefficient is a cost per acre. The objective is total production cost,
measured in dollars. The three lines group costs by land type.

## Production requirements

The model must produce the specified quantities of corn, wheat, and pasture
output:

$$
\begin{aligned}
80C_1+60C_2+40C_3 &= 7{,}500 &&\text{(corn)},\\
40W_1+35W_2+25W_3 &= 2{,}500 &&\text{(wheat)},\\
2.2P_1+2.0P_2+1.8P_3 &= 170 &&\text{(pasture)}.
\end{aligned}
$$

The coefficients are output per acre. Table 1 does not explicitly label the
physical output units, so we retain its numerical units. These are
**equalities**, as in the paper: the model produces the specified output rather
than choosing output in response to product prices.

## Land and cropping restrictions

Each land group initially contains 100 acres. Corn and wheat have maximum
acreages; pasture has a minimum acreage. All twelve restrictions are

$$
\begin{aligned}
C_1+W_1+P_1 &\le 100 &&\text{(land 1)},\\
C_1 &\le 60 &&\text{(corn 1 maximum)},\\
W_1 &\le 20 &&\text{(wheat 1 maximum)},\\
P_1 &\ge 20 &&\text{(pasture 1 minimum)},\\[5pt]
C_2+W_2+P_2 &\le 100 &&\text{(land 2)},\\
C_2 &\le 40 &&\text{(corn 2 maximum)},\\
W_2 &\le 30 &&\text{(wheat 2 maximum)},\\
P_2 &\ge 30 &&\text{(pasture 2 minimum)},\\[5pt]
C_3+W_3+P_3 &\le 100 &&\text{(land 3)},\\
C_3 &\le 20 &&\text{(corn 3 maximum)},\\
W_3 &\le 30 &&\text{(wheat 3 maximum)},\\
P_3 &\ge 30 &&\text{(pasture 3 minimum)}.
\end{aligned}
$$

Finally, the variable restrictions are

$$
C_1,W_1,P_1,C_2,W_2,P_2,C_3,W_3,P_3\ge0.
$$

Together, the objective, three production equalities, twelve land-use
inequalities, and nine nonnegativity restrictions define the full baseline LP.
The cropping bounds can represent physical suitability or imposed land-use
rules. Their economic justification matters because they help determine the
calculated value of development.

## Baseline allocation

Solving this model in R gives the following acreages and a minimized cost of
**\$`r money(baseline$cost)`**. The results and figure in these notes are
computed when the document renders.

```{r}
#| label: baseline-allocation
allocation <- matrix(baseline$acres, nrow = 3, byrow = TRUE)
knitr::kable(
  data.frame(`Land group` = 1:3, Corn = allocation[, 1],
             Wheat = allocation[, 2], Pasture = allocation[, 3],
             `Total used` = rowSums(allocation), check.names = FALSE),
  format = "latex", digits = 2, booktabs = TRUE, row.names = FALSE
)
```

Land group 2 uses all 100 acres, with corn and wheat at their upper bounds and
pasture at its lower bound. The economic question is how this whole production
system would adjust if more of that land package became available.

# The main insight: Value the complete resource package

## Specify the change before interpreting the shadow price

Let $L$ denote the endowment of land group 2. Keeping its characteristics in
the original proportions replaces its four bounds with

$$
\begin{aligned}
C_2+W_2+P_2&\le L,\\
C_2&\le0.4L,\\
W_2&\le0.3L,\\
P_2&\ge0.3L.
\end{aligned}
$$

At the baseline, $L=100$. An additional acre raises the four right-hand sides
by $1$, $0.4$, $0.3$, and $0.3$. The fractions apply to the **land endowment**;
the solver chooses the actual allocation. Raising the pasture minimum tightens
a restriction, so the package brings an additional obligation as well as
additional opportunities.

## Derive the weighted sum

Write $C^*(L)$ for minimized cost and $\lambda_i$ for its local change per unit
increase in right-hand side $b_i$. With the original inequality directions,
upper-bound resource prices are nonpositive; minimum-requirement prices are
nonnegative.

Within a range where the relevant marginal values remain valid,

$$
\Delta C^*=\lambda_L\Delta b_L+\lambda_C\Delta b_C
             +\lambda_W\Delta b_W+\lambda_P\Delta b_P.
$$

Substituting the coordinated changes gives

$$
\Delta C^*
=\left(\lambda_L+0.4\lambda_C+0.3\lambda_W+0.3\lambda_P\right)\Delta L.
$$

Therefore the marginal benefit of the package, measured as a cost saving, is

$$
\boxed{MV(L)=-\frac{dC^*(L)}{dL}
=-\left(\lambda_L+0.4\lambda_C+0.3\lambda_W+0.3\lambda_P\right).}
$$

This is a **weighted sum**, not a weighted average. The weights are the
right-hand-side changes per added acre; there is no division by their sum.
Using only $\lambda_L$ would value more total land while holding every cropping
bound fixed, which is a different economic experiment.

## A numerical interpretation

One valid set of multipliers for this numerical model, matching the rounded
values in the paper's Table 2, gives the following contributions:

| Constraint | Cost shadow price | Change per acre | Cost effect |
|:-----------|------------------:|----------------:|------------:|
| Total land 2 | $-1.14917$ | $1.0$ | $-1.14917$ |
| Corn maximum | $-1.55083$ | $0.4$ | $-0.62033$ |
| Wheat maximum | $0$ | $0.3$ | $0$ |
| Pasture minimum | $+1.44250$ | $0.3$ | $+0.43275$ |
| **Package** | | | **$-1.33675$** |

An additional acre saves approximately **\$`r money(baseline$mv)`** at the
margin. The extra pasture requirement offsets some of the gains from additional
land and permitted corn acreage. A direct re-solve at $L=100.001$ gives a cost
saving per added acre of `r sprintf("%.5f", finite_difference)`, confirming the
weighted-shadow-price calculation.

Individual multipliers need not be unique at this solution: the land and
cropping constraints bind together. A solver can report a different allocation
of shadow values across these rows while giving the same value for the
coordinated package. A binding constraint can also have a zero shadow price.

The paper's earlier displayed multipliers on printed page 5 imply about
\$1.34042, a slightly different value. These notes consistently use Table 1's
coefficients, which yield \$1.33675 and agree with the later range-analysis
tables after accounting for the transfer activity's cost.

# Representing the package with a transfer activity

Define a new decision variable $t\ge0$ as additional acres of the land group 2
package. Then $L=100+t$, and the four constraints become

$$
\begin{aligned}
C_2+W_2+P_2-t&\le100,\\
C_2-0.4t&\le40,\\
W_2-0.3t&\le30,\\
P_2-0.3t&\ge30.
\end{aligned}
$$

These are the coefficients in the paper's **TRANSFER 2** column. For example,
$t=5$ provides 105 total acres, permits up to 42 corn acres and 31.5 wheat
acres, and requires at least 31.5 pasture acres. It adds land to the modeled
resource base; it does not move land out of groups 1 or 3.

The other constraints and the nine production-cost coefficients stay as
specified above. The paper adds a cost of $0.01t$ to the objective and, in its
illustrative bounded experiment, adds the row $t\le1$. More generally, a bound
$t\le\bar t$ lets the analyst vary how much development is allowed.

If $k$ is development cost per acre, the expanded objective is production cost
plus $kt$. The local change in that objective per unit of transfer is

$$
k+\lambda_L+0.4\lambda_C+0.3\lambda_W+0.3\lambda_P
=k-MV(L).
$$

This links the transfer activity's reduced-cost calculation to the package
valuation. Expansion is attractive when its gross cost saving exceeds its
development cost. The paper's $0.01$ activity cost must be accounted for when
recovering gross resource value from its reported net activity value; reduced
cost signs also depend on the solver's reporting convention and active bound.

The transfer activity makes a coordinated change into a single explicit
choice. Its sensitivity range concerns the whole package. Separate
one-at-a-time ranges for land, corn, wheat, or pasture do not directly give
the allowable range for that coordinated change.

\newpage

# Derived demand for the land package

## Trace marginal value as the endowment changes

To construct the derived-demand schedule, vary $L$, update all four land group
2 bounds, and re-solve the production model. Keep costs, yields, the other land
groups, and output requirements fixed. At each solution calculate

$$
MV(L)=-\left(\lambda_L(L)+0.4\lambda_C(L)
                     +0.3\lambda_W(L)+0.3\lambda_P(L)\right).
$$

These are valuations of specified endowments. In particular, evaluating
$L<100$ describes a smaller endowment, rather than a negative value of the
nonnegative development variable $t$.

Within each optimal LP regime, cost is linear in $L$ and marginal value is
constant. When the limiting constraints or optimal production adjustments
change, the slope changes. The result is a stepwise marginal-value schedule.
At a corner, use the appropriate one-sided value for expansion or contraction.

The embedded R calculation uses `lpSolve`, audits feasibility and cost at every
solution, and checks the shadow-price prediction against small numerical
changes in $L$. It locates the plotted corners by intersecting adjacent linear
cost segments.

```{r}
#| label: fig-derived-demand
#| fig-cap: "Derived marginal value of the land group 2 package, computed from Table 1. All four land group 2 bounds change together. The dashed line marks the 100-acre baseline; dotted lines mark changes in marginal value. The negative segment reflects the rising pasture minimum."
#| fig-width: 6.2
#| fig-height: 3.8
#| out-width: "100%"
#| fig-align: center
#| fig-pos: H
#| dev: pdf
par(mar = c(4.2, 4.5, 1.0, 0.7), mgp = c(2.7, 0.8, 0),
    family = "serif", las = 1, cex = 0.95)
plot(range(edges), range(c(regime_mv, 0)) + c(-0.20, 0.28),
     type = "n", xlab = "Land group 2 endowment (acres)",
     ylab = "Marginal cost saving ($ per added acre)",
     xaxs = "i", yaxs = "i", bty = "l")
abline(h = 0, col = "gray55", lwd = 0.8)
abline(v = 100, lty = 2, col = "gray45")
abline(v = breaks, lty = 3, col = "gray65")
segments(head(edges, -1), regime_mv, tail(edges, -1), regime_mv,
         lwd = 2.4, col = "#174A66")
segments(breaks, head(regime_mv, -1), breaks, tail(regime_mv, -1),
         lwd = 1.1, lty = 3, col = "#174A66")
text((head(edges, -1) + tail(edges, -1)) / 2,
     regime_mv + 0.12, labels = sprintf("$%.2f", regime_mv), cex = 0.9)
points(100, baseline$mv, pch = 16, cex = 0.8)
text(100, par("usr")[4] - 0.08, "Baseline", pos = 2, cex = 0.8)
```

\newpage

## Interpret the steps and the development decision

The initial value of approximately \$`r money(regime_mv[1])` persists through
about **`r sprintf("%.2f", breaks[1])` acres**. At that point, the minimum
pasture acreage on land group 3 binds, changing the production adjustment
available to accommodate additional land group 2. The marginal benefit falls
to approximately **\$`r money(regime_mv[2])`**.

At about **`r sprintf("%.2f", breaks[2])` acres**, corn production on land
group 3 reaches zero and the marginal value falls again, becoming negative.
Additional land group 2 can no longer save costs by replacing corn on land
group 3. The model
now pays more to meet the same output requirements as $L$ rises. This is
possible because the package increases a compulsory pasture minimum:
larger endowments do not simply enlarge the feasible set. The paper's Figure 1
emphasizes the positive steps; the calculation here also displays the next
segment implied by its numerical model.

For an optional development project with nonnegative development costs, there
is no reason to purchase additions on the negative-value segment. With
constant cost $k$, compare the height of each step with $k$ and expand only
while the marginal saving exceeds that cost. For a finite project, total gross
benefit is the area under the marginal-value schedule:

$$
C^*(100)-C^*(100+t)=\int_{100}^{100+t}MV(L)\,dL.
$$

Subtract $kt$ to obtain net benefit. Multiplying the initial shadow value by
every added acre is valid only while the project stays within that initial
range. The value of the resource follows from both the **package being added**
and the **production adjustments available at each endowment**.

\newpage

# Model Lab Example

## Question and experimental design

**What is the production-cost penalty from applying the land-use restrictions
to newly developed group 2 acreage, and how does that penalty change with
project size?**

Compare two ways of adding $t$ acres of group 2 land. Restricted expansion
brings the paper's corn and wheat maxima and pasture minimum. Unrestricted
expansion permits any mix of the three activities on the additional land.
Both scenarios retain the original restrictions on all existing land, the
same yields and costs, and the same output requirements. This isolates the
value of flexibility on the new acreage.

Increasing the total-land bound alone while holding the corn and wheat bounds
fixed would not implement unrestricted expansion: those crop bounds would
still apply to the additional acreage. Instead, distinguish existing and new
land explicitly.

Let $z_C,z_W,z_P\ge0$ denote acres of corn, wheat, and pasture on the **new**
group 2 land. The original nine variables describe only existing land. Retain
all twelve original land-use inequalities and all nine original nonnegativity
restrictions. Replace the output equalities with

$$
\begin{aligned}
80C_1+60C_2+40C_3+60z_C &=7{,}500,\\
40W_1+35W_2+25W_3+35z_W &=2{,}500,\\
2.2P_1+2.0P_2+1.8P_3+2.0z_P &=170.
\end{aligned}
$$

If $C_{\mathrm{old}}(x)$ denotes the fully parameterized production-cost
objective given above, both scenarios minimize

$$
C_{\mathrm{old}}(x)+151.20z_C+93.80z_W+84.16z_P.
$$

The **unrestricted** scenario adds only

$$
z_C+z_W+z_P\le t,\qquad z_C,z_W,z_P\ge0.
$$

The **restricted** scenario also imposes

$$
z_C\le0.4t,\qquad z_W\le0.3t,\qquad z_P\ge0.3t.
$$

These are maxima and a minimum, not three required exact shares. The
restricted formulation is equivalent, in aggregate, to expanding the original
group 2 package to $100+t$ acres. Because old and new group 2 land have the
same technology, labels on otherwise interchangeable acres can be nonunique;
the minimized cost is the quantity we use for valuation.

## Define the value of removing the restrictions

Let $C_R(t)$ and $C_U(t)$ be minimized production costs in the restricted and
unrestricted scenarios, and let $C_0=C_R(0)=C_U(0)$ be baseline cost. Gross
benefits of development are

$$
B_R(t)=C_0-C_R(t),\qquad B_U(t)=C_0-C_U(t).
$$

The total value of flexibility on the added acreage is

$$
\boxed{V(t)=C_R(t)-C_U(t)=B_U(t)-B_R(t).}
$$

Every allocation feasible under restricted expansion is also feasible under
unrestricted expansion. Therefore $C_U(t)\le C_R(t)$ and **$V(t)\ge0$** wherever
both problems are feasible. The average value per added acre is $V(t)/t$ for
$t>0$; it is distinct from the marginal value $V'(t)$.

No development cost is included in this comparison. An identical cost $kt$
in both scenarios would cancel from $V(t)$. The result is the production
system's value of removing the restrictions. Any benefits of the restrictions
themselves, such as erosion reduction, would require separate representation.

Define the marginal benefits of acreage as

$$
MV_R(t)=-C_R'(t),\qquad MV_U(t)=-C_U'(t).
$$

For the new-land constraints, denote cost shadow prices by $\mu$. As in the
paper, restricted expansion changes four bounds together, whereas
unrestricted expansion changes only the new-land availability bound:

$$
\begin{aligned}
MV_R(t)&=-\left(\mu_{L,R}+0.4\mu_{C,R}
                         +0.3\mu_{W,R}+0.3\mu_{P,R}\right),\\
MV_U(t)&=-\mu_{L,U}.
\end{aligned}
$$

Thus, away from corners,

$$
\boxed{V'(t)=MV_U(t)-MV_R(t),\qquad
V(t)=\int_0^t\left[MV_U(s)-MV_R(s)\right]ds.}
$$

Use one-sided slopes at corners and at $t=0$. In particular, a solver may
return a nonunique dual solution when all new-acreage variables are fixed to
zero; a small positive expansion identifies the relevant right-hand slope.

```{r}
#| label: lab-calculations
#| include: false
library(ggplot2)

solve_expansion <- function(t, restricted = FALSE) {
  stopifnot(length(t) == 1L, is.finite(t), t >= 0)
  new_columns <- matrix(0, nrow(A), 3)
  new_columns[1:3, ] <- diag(c(60, 35, 2))
  lab_A <- rbind(cbind(A, new_columns), c(rep(0, 9), 1, 1, 1))
  lab_dir <- c(directions, "<=")
  lab_rhs <- c(7500, 2500, 170, 100, 60, 20, 20,
               100, 40, 30, 30, 100, 20, 30, 30, t)
  rhs_change <- 1
  if (restricted) {
    lab_A <- rbind(lab_A, cbind(matrix(0, 3, 9), diag(3)))
    lab_dir <- c(lab_dir, "<=", "<=", ">=")
    lab_rhs <- c(lab_rhs, 0.4 * t, 0.3 * t, 0.3 * t)
    rhs_change <- c(1, 0.4, 0.3, 0.3)
  }
  lab_cost <- c(cost, cost[4:6])
  fit <- lp("min", lab_cost, lab_A, lab_dir, lab_rhs,
            compute.sens = TRUE)
  if (fit$status != 0) stop("Expansion problem not optimal at t = ", t)
  residual <- drop(lab_A %*% fit$solution) - lab_rhs
  stopifnot(all(fit$solution >= -1e-7),
            all(abs(residual[lab_dir == "="]) < 1e-6),
            all(residual[lab_dir == "<="] < 1e-6),
            all(residual[lab_dir == ">="] > -1e-6),
            abs(sum(lab_cost * fit$solution) - fit$objval) < 1e-6)
  list(cost = fit$objval, acres = fit$solution,
       mv = -sum(fit$duals[16:nrow(lab_A)] * rhs_change))
}

# Sample inside intervals so t=0 and known corners do not select ambiguous
# endpoint duals. Intersect adjacent linear cost segments for exact corners.
lab_grid <- seq(0, 20, by = 0.025)
lab_midpoints <- (head(lab_grid, -1) + tail(lab_grid, -1)) / 2
get_lab_regimes <- function(restricted) {
  fits <- lapply(lab_midpoints, solve_expansion, restricted = restricted)
  mv <- vapply(fits, function(s) s$mv, numeric(1))
  mv[abs(mv) < 1e-7] <- 0
  starts <- c(1L, which(abs(diff(mv)) > 1e-6) + 1L)
  slopes <- mv[starts]
  intercept <- vapply(fits[starts], function(s) s$cost, numeric(1)) +
    slopes * lab_midpoints[starts]
  corners <- diff(intercept) / diff(slopes)
  bounds <- c(0, corners, 20)
  stopifnot(all(diff(bounds) > 0))
  segments <- data.frame(t_start = head(bounds, -1),
                         t_end = tail(bounds, -1), mv = slopes,
                         intercept = intercept)
  for (j in seq_len(nrow(segments))) {
    t <- mean(c(segments$t_start[j], segments$t_end[j]))
    fd <- (solve_expansion(t, restricted)$cost -
           solve_expansion(t + epsilon, restricted)$cost) / epsilon
    stopifnot(abs(fd - segments$mv[j]) < 1e-5)
  }
  segments
}
restricted_regimes <- get_lab_regimes(TRUE)
unrestricted_regimes <- get_lab_regimes(FALSE)
regime_cost <- function(t, segments) {
  i <- pmin(findInterval(t, segments$t_start), nrow(segments))
  segments$intercept[i] - segments$mv[i] * t
}

# Independently verify the piecewise curves against grid solutions, confirm
# feasible-set dominance, and reproduce the paper's aggregate formulation.
for (t in lab_grid) {
  r <- solve_expansion(t, TRUE)
  u <- solve_expansion(t, FALSE)
  stopifnot(abs(r$cost - regime_cost(t, restricted_regimes)) < 1e-5,
            abs(u$cost - regime_cost(t, unrestricted_regimes)) < 1e-5,
            abs(r$cost - solve_land(100 + t)$cost) < 1e-5,
            u$cost <= r$cost + 1e-6)
}
stopifnot(abs(solve_expansion(0, TRUE)$cost - baseline$cost) < 1e-6,
          abs(solve_expansion(0, FALSE)$cost - baseline$cost) < 1e-6)

example_t <- c(0, 1, 5, 10, 12.5, 20)
examples <- data.frame(t = example_t,
                      restricted = regime_cost(example_t, restricted_regimes),
                      unrestricted = regime_cost(example_t, unrestricted_regimes))
examples$value <- examples$restricted - examples$unrestricted
examples$value[abs(examples$value) < 1e-7] <- 0
ten <- examples[examples$t == 10, ]
initial_R <- restricted_regimes$mv[1]
initial_U <- unrestricted_regimes$mv[1]
initial_gap <- initial_U - initial_R
unrestricted_plateau <- tail(unrestricted_regimes$t_start, 1)

# Plot only actual segment endpoints; lines therefore retain the LP corners.
plot_t <- sort(unique(c(restricted_regimes$t_start,
                        unrestricted_regimes$t_start, 20)))
plot_R <- baseline$cost - regime_cost(plot_t, restricted_regimes)
plot_U <- baseline$cost - regime_cost(plot_t, unrestricted_regimes)
benefit_data <- rbind(
  data.frame(t = plot_t, benefit = plot_R, scenario = "Restricted"),
  data.frame(t = plot_t, benefit = plot_U, scenario = "Unrestricted")
)
value_data <- data.frame(t = plot_t, value = plot_U - plot_R)
benefit_band <- data.frame(t = plot_t, lower = plot_R, upper = plot_U)
scenario_colors <- c(Restricted = "#A34A28", Unrestricted = "#174A66")
scenario_lines <- c(Restricted = "dashed", Unrestricted = "solid")
lab_theme <- theme_minimal(base_size = 10, base_family = "serif") +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom",
        legend.title = element_blank(), plot.margin = margin(6, 10, 6, 6),
        strip.text = element_text(face = "bold", hjust = 0))

# The area between marginal curves equals the directly calculated V(t).
interval_mid <- (head(plot_t, -1) + tail(plot_t, -1)) / 2
get_mv <- function(t, segments) {
  segments$mv[pmin(findInterval(t, segments$t_start), nrow(segments))]
}
interval_R <- get_mv(interval_mid, restricted_regimes)
interval_U <- get_mv(interval_mid, unrestricted_regimes)
integrated_value <- c(0, cumsum((interval_U - interval_R) * diff(plot_t)))
stopifnot(max(abs(integrated_value - value_data$value)) < 1e-5)
```

\newpage

## Calculate the value of flexibility

The calculation solves both twelve-variable LPs at each expansion size. It
checks feasibility, reconstructs costs, verifies $C_U(t)\le C_R(t)$, and confirms
that restricted expansion reproduces the earlier aggregate model. The table
reports dollars over the same accounting period as the original model.

```{r}
#| label: lab-value-table
knitr::kable(
  data.frame(
    `Added acres` = sprintf("%g", examples$t),
    `Restricted cost` = money(examples$restricted),
    `Unrestricted cost` = money(examples$unrestricted),
    `Flexibility value` = money(examples$value),
    `Value per acre` = ifelse(examples$t == 0, "--",
                             money(examples$value / examples$t)),
    check.names = FALSE
  ), format = "latex", booktabs = TRUE, row.names = FALSE,
  align = c("r", "r", "r", "r", "r")
)
```

For a **10-acre project**, restricted expansion saves
\$`r money(baseline$cost - ten$restricted)`, while unrestricted expansion
saves \$`r money(baseline$cost - ten$unrestricted)`. Removing the restrictions
is therefore worth **\$`r money(ten$value)` in total**, or
**\$`r money(ten$value / 10)` per added acre** on average. This is the maximum
additional project cost justified by flexibility alone within this model.

```{r}
#| label: fig-lab-development-benefits
#| fig-cap: "Production-cost savings from adding group 2 land. The vertical gap, shaded between the curves, is the total value of removing restrictions at a given project size. The baseline is the same in both scenarios."
#| fig-width: 6.2
#| fig-height: 3.6
#| out-width: "100%"
#| fig-align: center
#| fig-pos: H
#| dev: pdf
ggplot(benefit_data, aes(t, benefit, color = scenario, linetype = scenario)) +
  geom_ribbon(data = benefit_band,
              aes(x = t, ymin = lower, ymax = upper),
              inherit.aes = FALSE, fill = "#DCE8EC", alpha = 0.8) +
  geom_line(linewidth = 0.85) +
  annotate("segment", x = 10, xend = 10,
           y = baseline$cost - ten$restricted,
           yend = baseline$cost - ten$unrestricted,
           linewidth = 0.65, color = "gray25") +
  annotate("text", x = 10.4,
           y = baseline$cost - (ten$restricted + ten$unrestricted) / 2,
           label = paste0("$", money(ten$value), " at 10 acres"),
           hjust = 0, size = 3.1, family = "serif") +
  scale_color_manual(values = scenario_colors) +
  scale_linetype_manual(values = scenario_lines) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Additional group 2 land (acres)",
       y = "Total production-cost saving ($)") + lab_theme
```

The unrestricted curve levels off after approximately
**`r sprintf("%.2f", unrestricted_plateau)` added acres**. Beyond that point,
more available acreage creates no further production-cost saving. Restricted
benefits peak at **12.50 added acres** and then decline as the growing pasture
minimum becomes costly. Common development costs would be subtracted from
both benefit curves when choosing project size.

\newpage

## Explain the marginal values and their changing gap

Initially, one unrestricted acre can grow corn and replace $60/40=1.5$ acres
of corn on group 3. The saving is

$$
MV_U(0^+)=1.5(102.60)-151.20=\$2.70\text{ per added acre}.
$$

This adjustment lasts for five added acres, enough to replace the original
7.5 acres of group 3 corn. Restricted expansion initially saves only
\$`r sprintf("%.5f", initial_R)` per acre, because the corn maximum expands
more slowly and the pasture requirement grows. Thus

$$
V'(0^+)=`r sprintf("%.5f", initial_U)`-`r sprintf("%.5f", initial_R)`
=\$`r sprintf("%.5f", initial_gap)`\text{ per added acre}.
$$

The initial value of flexibility is about \$`r money(initial_gap)` per acre,
but this rate cannot be multiplied by every acre of a large project.

```{r}
#| label: fig-lab-marginal-and-total-flexibility
#| fig-cap: "Upper panel: marginal values of added acreage. Lower panel: total value of removing its restrictions. Total flexibility value rises where the unrestricted marginal value is higher and falls where it is lower. Dotted guides mark unrestricted breakpoints; total flexibility value remains nonnegative."
#| fig-width: 6.2
#| fig-height: 4.8
#| out-width: "100%"
#| fig-align: center
#| fig-pos: H
#| dev: pdf
marginal_data <- rbind(
  transform(restricted_regimes, scenario = "Restricted"),
  transform(unrestricted_regimes, scenario = "Unrestricted")
)
panel_levels <- c("Marginal cost saving ($ per added acre)",
                  "Total value of flexibility ($)")
marginal_data$panel <- factor(panel_levels[1], levels = panel_levels)
value_data$panel <- factor(panel_levels[2], levels = panel_levels)
ggplot() +
  geom_hline(yintercept = 0, color = "gray65", linewidth = 0.3) +
  geom_vline(xintercept = unrestricted_regimes$t_start[-1],
             color = "gray70", linetype = "dotted", linewidth = 0.4) +
  geom_segment(data = marginal_data,
               aes(x = t_start, xend = t_end, y = mv, yend = mv,
                   color = scenario, linetype = scenario), linewidth = 0.9) +
  geom_line(data = value_data, aes(t, value),
            color = "gray15", linewidth = 0.85) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  scale_color_manual(values = scenario_colors) +
  scale_linetype_manual(values = scenario_lines) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Additional group 2 land (acres)", y = NULL) + lab_theme
```

## What the experiment teaches

**A nonnegative total value need not increase at every project size.**
Unrestricted expansion always costs no more than restricted expansion at the
same $t$. Nevertheless, the marginal-value curves cross. Between 5 and about
5.52 added acres, unrestricted corn substitution has ended, while restricted
expansion still earns its initial marginal benefit. Between about
`r sprintf("%.2f", unrestricted_plateau)` and 12.50 acres, unrestricted benefits
are flat while restricted benefits continue rising. In both intervals,
$V'(t)<0$: the restricted scenario partially catches up. That does not imply
the restrictions have become beneficial.

After 12.50 acres, the restricted scenario's pasture obligation raises costs,
while unrestricted cost stays constant. Consequently the value of removing
the restrictions rises again. The integral of the marginal-value gap matches
the direct cost difference; the R calculation checks both ways of computing
$V(t)$.

**Project size and the decision being valued must be explicit.** These curves
compare two specified projects of equal size. An unrestricted project may
leave acreage unused. A restricted project still incurs its pasture minimum
on all developed acres, even if it would prefer a smaller project. The plotted
comparison stops at $t=20$: beyond that, the three pasture minima alone imply
more than the required 170 units of pasture output, making restricted
expansion infeasible under the paper's equality constraint.

For a Model Lab presentation, the central explanation is the sequence from
the counterfactual to its consequences: define which land is freed from which
rules, predict the initial marginal effect, solve and verify both models, and
use changing allocations and binding constraints to explain the difference
between the two value curves.
