---
title: "Model Lab: The Value of Land-Use Flexibility"
format:
  revealjs:
    theme: simple
    title-slide: false
    slide-number: true
    transition: fade
    center: false
    width: 1280
    height: 720
    margin: 0.04
    embed-resources: true
    toc: false
    number-sections: false
    include-in-header:
      text: |
        <style>
        .reveal { font-size: 32px; }
        .reveal h2 { font-size: 1.45em; margin-bottom: 0.6em; }
        .reveal h3 { font-size: 1.05em; margin-top: 0.5em; }
        .reveal p, .reveal li { line-height: 1.3; }
        .reveal .shared-assumptions { font-size: 0.85em; margin-top: 1.1em; }
        .reveal .valuation { text-align: center; font-size: 1.18em; margin: 0.8em 0; }
        .reveal .result { font-size: 0.9em; text-align: center; }
        .reveal .cell-output-display img { margin: 0 auto; }
        </style>
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: slide-calculations
#| include: false
# Reuse the model and audited calculations from the neighboring notes.
# Rendering this QMD requires Stewart_Notes.qmd, lpSolve, ggplot2, and knitr.
# The rendered HTML embeds its figure and presentation resources.
notes_source <- readLines("Stewart_Notes.qmd", warn = FALSE)
notes_chunk <- function(label) {
  start <- which(trimws(notes_source) == paste0("#| label: ", label))
  if (length(start) != 1L) stop("Expected one notes chunk named: ", label)
  closing <- which(seq_along(notes_source) > start &
                     trimws(notes_source) == "```")
  if (!length(closing)) stop("Missing closing fence for: ", label)
  notes_source[seq.int(start + 1L, min(closing) - 1L)]
}
for (label in c("setup-model", "lab-calculations")) {
  eval(parse(text = notes_chunk(label)), envir = knitr::knit_global())
}

```

## What is the value of unrestricted new acreage?

Add *t* acres of **group 2 land** and compare two versions of the same project.

:::: {.columns}
::: {.column width="50%"}
### Restricted expansion

On the additional acreage:

- Corn: at most 40%.
- Wheat: at most 30%.
- Pasture: at least 30%.
:::

::: {.column width="50%"}
### Unrestricted expansion

Any combination of corn, wheat, and pasture on the new land.

Total use cannot exceed the additional acreage available.
:::
::::

::: {.shared-assumptions}
**Existing land keeps its restrictions.** Both projects use the same yields,
production costs, and output requirements.
:::

::: {.notes}
This extends Stewart and Greenhalgh (1977). Only the treatment of newly added
group 2 acreage changes. Restricted shares are upper bounds for corn and wheat
and a lower bound for pasture, rather than three exact required shares.
Increasing the total-land bound alone while retaining the original aggregate
crop caps would not make the new land unrestricted. The notes implement
separate activity variables for the additional acres.
:::

## Policy relevance of land-use flexibility

A development program may require pasture or limit cropping on new land.
What production benefit does the program give up by attaching those rules?

::: {.valuation}
<i>V</i>(<i>t</i>) = <i>C</i><sub>R</sub>(<i>t</i>) − <i>C</i><sub>U</sub>(<i>t</i>)
= <i>B</i><sub>U</sub>(<i>t</i>) − <i>B</i><sub>R</sub>(<i>t</i>)
:::

Here, **C** is minimized production cost and **B** is the saving from expansion.
**V** measures the value of removing the restrictions for a project of size *t*.

- Compare that production cost with the restrictions' conservation benefits.
- Evaluate the proposed project size: the value per acre changes with acreage.

::: {.notes}
R denotes restricted expansion and U denotes unrestricted expansion. Both
benefits use the same baseline production cost. Feasible-set inclusion implies
V(t) is nonnegative, but it need not rise at every project size. Environmental
benefits are outside this production model and must be evaluated separately.
The paper's coefficients are illustrative, so these dollars describe the
example rather than an estimate for a current real-world policy.
Source: Stewart_Notes.qmd, Model Lab Example.
:::

## Flexibility is worth \$`r money(ten$value)` for 10 added acres

```{r}
#| label: flexibility-gap-plot
#| fig-width: 10
#| fig-height: 4.2
#| out-width: "94%"
#| fig-align: center
#| fig-alt: "Restricted and unrestricted production-cost savings as group 2 acreage expands. The shaded gap is the value of flexibility. At 10 additional acres, the gap is 7.16 dollars."
#| dev: svg
# Recreate Figure 2 with the same computed values and larger slide labels.
ggplot(benefit_data, aes(t, benefit, color = scenario, linetype = scenario)) +
  geom_ribbon(data = benefit_band,
              aes(x = t, ymin = lower, ymax = upper),
              inherit.aes = FALSE, fill = "#DCE8EC", alpha = 0.8) +
  geom_line(linewidth = 1.15) +
  annotate("text", x = 19.8,
           y = baseline$cost - regime_cost(19.8, unrestricted_regimes) + 0.8,
           label = "Unrestricted", color = scenario_colors[["Unrestricted"]],
           hjust = 1, size = 5.6, fontface = "bold" ) +
  annotate("text", x = 19.8,
           y = baseline$cost - regime_cost(19.8, restricted_regimes) - 0.9,
           label = "Restricted", color = scenario_colors[["Restricted"]],
           hjust = 1, size = 5.6, fontface = "bold" ) +
  annotate("segment", x = 10, xend = 10,
           y = baseline$cost - ten$restricted,
           yend = baseline$cost - ten$unrestricted,
           linewidth = 0.8, color = "gray25") +
  annotate("text", x = 10.45,
           y = baseline$cost - (ten$restricted + ten$unrestricted) / 2,
           label = paste0("$", money(ten$value), " value of flexibility"),
           hjust = 0, size = 5) +
  scale_color_manual(values = scenario_colors) +
  scale_linetype_manual(values = scenario_lines) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Additional group 2 land (acres)",
       y = "Total cost saving ($)") +
  theme_minimal(base_size = 17) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "none",
        plot.margin = margin(5, 12, 0, 5))
```

::: {.result}
At 10 acres, total savings are **\$`r money(baseline$cost - ten$unrestricted)` unrestricted**
versus **\$`r money(baseline$cost - ten$restricted)` restricted**.
The **\$`r money(ten$value)` difference** is the value of flexibility.
:::

::: {.notes}
This reproduces Figure 2 from the notes. At any project size, the vertical gap
between the cost-saving curves is the total value of flexibility. At ten acres,
the difference is approximately $7.16, or $0.72 per added acre on average.
The unrestricted benefit levels off after about 8.38 acres. Restricted
benefits peak at 12.5 acres and then decline as the pasture minimum grows.
The shrinking and widening gap shows why the initial marginal value of
flexibility cannot be applied to every acre of a larger project. All monetary
values use the original model's accounting period, not capitalized land prices.
:::

## Calculating the cost-savings curves

**Shadow prices apply within a segment. We recomputed them as marginal values changed.**

1. **Re-solve both LPs** over 0–20 added acres using a grid spaced 0.025 acres apart.
2. **Record minimized costs and shadow prices.** Use the weighted sum for restricted
   expansion and the new-land shadow price for unrestricted expansion.
3. **Construct each savings curve as baseline cost minus optimized cost.** Locate
   corners by intersecting adjacent cost segments, then plot their endpoints.

::: {.shared-assumptions}
**Breakpoints in added acreage:** restricted at
**`r sprintf("%.2f", restricted_regimes$t_start[2])`** and
**`r sprintf("%.2f", restricted_regimes$t_start[3])`** acres; unrestricted at
**`r sprintf("%.2f", unrestricted_regimes$t_start[2])`** and
**`r sprintf("%.2f", unrestricted_regimes$t_start[3])`** acres.
One baseline shadow price cannot describe the entire curve.
:::

::: {.notes}
The calculation first solves both models at the midpoints of a grid spaced
0.025 acres apart. At each solution it records minimized cost and the marginal
cost saving implied by the duals. For restricted expansion this is minus the
weighted sum with weights 1, 0.4, 0.3, and 0.3. For unrestricted expansion it is
minus the shadow price on the new-land availability constraint.

Consecutive solutions with the same marginal saving define candidate linear
cost segments. Intersections of adjacent cost lines locate the breakpoints
more precisely than the grid spacing. The code verifies the resulting curves
against separate solves at all grid endpoints and verifies marginal values
with small changes in acreage inside each segment. It plots savings at the
union of the two scenarios' breakpoints, joined by straight lines.

Within a segment, its shadow value gives the exact linear change in optimized
cost. Beyond its valid range, the optimization must supply a new marginal
value. A parametric LP method could move directly between breakpoints; this
implementation uses repeated solves to discover and check them. At t=0 and
other corners, duals can be nonunique, so the calculation uses interior points
and one-sided comparisons rather than extrapolating an arbitrary endpoint dual.
Source: Stewart_Notes.qmd, lab-calculations chunk.
:::
