---
title: "Schlachtberger et al.: Formulating an Electricity Planning LP"
subtitle: "From the planning problem to the value of transmission"
format:
  pdf:
    papersize: letter
    geometry: margin=0.8in
    fontsize: 11pt
    number-sections: false
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
# Published scenario results, Table 2. This document does not re-solve
# the European model. The calculations and plot use these four observations.
volume <- c(0, 31.25, 125, 285.70)
cost <- c(84.1, 75.7, 67.5, 64.8)
benefit_share <- (cost[1] - cost[3]) / (cost[1] - cost[4])
volume_share <- volume[3] / volume[4]
slopes <- diff(cost) / diff(volume)
stopifnot(all(diff(volume) > 0), all(diff(cost) < 0),
          all(diff(slopes) > 0),
          abs(benefit_share - 0.8601036) < 1e-6)
```

## 1. Start with the planning problem

**Reading:** Schlachtberger, D. P., Brown, T., Schramm, S., and Greiner, M.
(2017). "The benefits of cooperation in a highly renewable European
electricity network." *Energy* 134: 469--481.
[Public manuscript](https://arxiv.org/abs/1704.05492).

- European countries must supply electricity in every hour while sharply
reducing carbon emissions. 
- Wind and solar are low-emission sources, but their available output varies with weather and time of day. 
- Three ways to cover a local shortfall are to generate dispatchable electricity, discharge
stored energy, or import electricity from another country. 
- Each requires infrastructure and incurs costs or uses a limited resource.

**The planning question:** What generation, storage, and international
transmission capacities should be built, and how should they operate, to
meet demand at minimum annual cost subject to an emissions limit?

The paper represents 30 countries with one node per country, hourly demand
and weather from 2011, and a CO$_2$ cap 95% below 1990 electricity-sector
emissions. The network connects countries along the existing topology.
The model chooses capacities and operations together. 

### Separate choices from information

Demand, renewable availability, costs, efficiencies, connection lengths,
resource potentials, and policy caps are **given**. Infrastructure capacities,
hourly production, storage operation, and electricity flows are **chosen**.

Existing hydro and pumped-hydro capacities are fixed in the paper. New
hydro capacity is not an investment choice.

### Define variables before writing equations

Use $n$ for countries, $s$ for **generation and storage technologies**,
$t$ for hours, and $\ell$ for international connections, as in the paper.

| Variable | Meaning | Units |
|:--|:--|:--|
| $G_{n,s}$ | Generation or storage power capacity | MW |
| $g_{n,s,t}$ | Generation or net storage dispatch | MW |
| $soc_{n,s,t}$ | Storage state of charge | MWh |
| $F_\ell,f_{\ell,t}$ | Line capacity and signed flow | MW |

Storage uses the same $G_{n,s}$ and $g_{n,s,t}$ notation as generation.
For storage, positive dispatch supplies the grid; negative dispatch charges it.

30 countries, 4 technologies, and 8,760 hours give 1,051,200 $g$ variables.

\newpage

## 2. Translate electricity supply into constraints

### "Demand must be met in every country in every hour"

Let $d_{n,t}$ be demand. Define $K_{n\ell}=+1$ where positive flow leaves
country $n$, $-1$ where it enters, and zero otherwise. Then

$$
\underbrace{\sum_s g_{n,s,t}}_{\text{generation and net storage dispatch}}
-d_{n,t}
=\underbrace{\sum_\ell K_{n\ell}f_{\ell,t}}_{\text{net exports}}
\qquad \forall n,t.
$$

Equation (2). Charging enters $g_{n,s,t}$ negatively;
discharging enters positively. Exports reduce supply available for domestic
demand. Conservation requires equality.

**Check the signs.** If country A generates 80 MW, serves 60 MW of demand,
and charges storage at 5 MW, it can export 15 MW. At country B, that same
flow enters with the opposite sign. Summing balances over countries cancels
international flows: transmission redistributes electricity rather than
creating it.

$$
80-60-5=15
$$

**Why not impose just an annual balance?** Annual production can equal
annual demand while production falls short on a particular evening. Every
hour must be feasible, including hours of high demand and little wind.

### "A generator cannot produce more than its available capacity"

Let $\bar g_{n,s,t}$ denote available output per MW of installed capacity:

$$
0\le g_{n,s,t}\le \bar g_{n,s,t}G_{n,s}\qquad\forall n,s,t.
$$

- These bounds apply to generators, not signed storage dispatch.
- For gas turbines, use $\bar g=1$ in this formulation. 
- For wind and solar,
availability is determined by weather.
- Output may be below the available amount. The inequality permits **curtailment** when additional renewable output cannot be used or stored.

### "Suitable locations are limited"

$$
0\le G_{n,s}\le G^{\max}_{n,s}.
$$

- The upper bound represents geographic installation potential. 
- Omit a finite upper bound where the model assumes no such limit; fix capacity for existing hydro facilities. 
- These restrictions apply to installed capacity, while
weather availability restricts its hourly use.



\newpage

## 3. Translate storage, networks, and policy limits

### "Electricity can be shifted across hours, with losses"

The paper denotes charging and discharging efficiencies by $\eta_1$ and
$\eta_2$. Equation (6) is

$$
\begin{aligned}
soc_{n,s,t}={}&soc_{n,s,t-1}
+\eta_1 g_{n,s,t,\mathrm{charge}}
-\eta_2^{-1}g_{n,s,t,\mathrm{discharge}}\\
&+g_{n,s,t,\mathrm{inflow}}-g_{n,s,t,\mathrm{spillage}}.
\end{aligned}
$$

- Time steps are one hour: each power term implicitly multiplies one hour to
give energy. 
- Charging at 10 MW with $\eta_1=0.9$ adds 9 MWh; delivering
9 MWh with $\eta_2=0.9$ removes 10 MWh from storage.
- Loss or leakage in and out of storage. Like a tax



$$
0\le soc_{n,s,t}\le E_{n,s}=h_{s,\max}G_{n,s},\qquad
soc_{n,s,0}=soc_{n,s,T}.
$$

- The fixed duration $h_{s,\max}$ is 6 hours for batteries and 168 hours for
hydrogen. 
- A 100 MW battery therefore has 600 MWh of energy capacity.
- The cyclic condition prevents a free annual withdrawal from the initial stock.


### "Each connection has a carrying capacity"

$$
-F_\ell\le f_{\ell,t}\le F_\ell\qquad\forall\ell,t,
\qquad F_\ell\ge0.
$$

- Flows can reverse direction. 
- Capacity cannot be negative. 
- Two linear inequalities implement the bound $|f_{\ell,t}|\le F_\ell$.

### "Total international transmission development is restricted"

$$
\sum_\ell l_\ell F_\ell\le \mathrm{CAP}_{LV}.
$$

- Line length $l_\ell$ is fixed. The cap $\mathrm{CAP}_{LV}$ is measured in MWkm; a 1,000 MW connection spanning 500 km uses 500,000 MWkm. 
- This is a physical development limit, separate from the construction costs in the objective.

### "Total carbon emissions cannot exceed a budget"

$$
\sum_{n,s,t}\frac{e_s}{\eta_s}g_{n,s,t}\le \mathrm{CAP}_{\mathrm{CO}_2}.
$$

- Here $e_s$ is tonnes of CO$_2$ per MWh of fuel and $\eta_s$ is generation
efficiency. 
- Electrical output divided by efficiency gives fuel use. For non-emitting generation, $e_s=0$. 
- The carbon budget limits annual fuel use; it does not directly specify gas capacity or a renewable capacity share.

\newpage

## 4. Assemble the LP

### "Choose the least-cost feasible electricity system"

- The paper uses $c_{n,s}$ for annualized generation and storage capacity
costs, 
- $c_\ell$ for annualized line costs, and 
- $o_{n,s}$ for variable dispatch costs. 
- Storage capacity costs include energy capacity through the fixed ratio $h_{s,\max}$ i.e., $E_{n,s}=h_{s,\max}G_{n,s}$.

Using the paper's notation, the objective and constraints are

$$
\begin{aligned}
\min_{G,F,g,f,soc}\quad
&\sum_{n,s}c_{n,s}G_{n,s}+\sum_\ell c_\ell F_\ell
+\sum_{n,s,t}o_{n,s}g_{n,s,t}\\
\text{subject to}\quad
&\sum_s g_{n,s,t}-d_{n,t}=\sum_\ell K_{n\ell}f_{\ell,t} \quad \text{(demand)},\\
&0\le g_{n,s,t}\le\bar g_{n,s,t}G_{n,s} 
\quad\text{(generators)},\\
&0\le G_{n,s}\le G^{\max}_{n,s},\\
&soc_{n,s,t}=soc_{n,s,t-1}
+\eta_1g_{n,s,t,\mathrm{charge}}
-\eta_2^{-1}g_{n,s,t,\mathrm{discharge}}\\
&\hspace{2cm}+g_{n,s,t,\mathrm{inflow}}-g_{n,s,t,\mathrm{spillage}},\\
&g_{n,s,t}=g_{n,s,t,\mathrm{discharge}}-g_{n,s,t,\mathrm{charge}}
\quad\text{(storage)},\\
&0\le g_{n,s,t,\mathrm{charge}},g_{n,s,t,\mathrm{discharge}}\le G_{n,s},\\
&0\le soc_{n,s,t}\le h_{s,\max}G_{n,s},\qquad soc_{n,s,0}=soc_{n,s,T},\\
&-F_\ell\le f_{\ell,t}\le F_\ell,\qquad F_\ell\ge0. \quad \text{(transmission capacity)},\\
&\sum_\ell l_\ell F_\ell\le\mathrm{CAP}_{LV} \quad  \text{(line capacity)},\\
&\sum_{n,s,t}(e_s/\eta_s)g_{n,s,t}\le\mathrm{CAP}_{\mathrm{CO}_2} \quad \text{(carbon budget)}.
\end{aligned}
$$

Apply generator and storage rows only to the corresponding technologies.
Storage component flows are included in the choices summarized by $g$;
inflow is given and spillage is nonnegative. Fix existing hydro capacities,
use $\bar g=1$ for gas, and omit capacity upper bounds where unrestricted.
All sums over time use the paper's implicit one-hour steps.

### Check the formulation before solving

1. **Units:** Each balance is in MW; each storage balance is in MWh;
   emissions are in tonnes; every objective term is in euros per year.
2. **Signs and domains:** Charging and exports use local electricity.
   Generator output, stocks, and capacities are nonnegative; net storage
   dispatch and line flows are signed.
3. **Linearity:** Costs, availability, efficiencies, durations, and lengths
   are parameters. There are no products of decision variables.
4. **Completeness:** Capacity constraints link investments to hourly operation.
   The storage recurrence links hours. The network links countries.

Stacking variables gives $\min c^\top x$ subject to linear equalities and
inequalities. If a solver requires nonnegative variables, write
$f=f^+-f^-$, with $f^+,f^-\ge0$. 

> **Question:** What becomes possible if we accidentally omit $g_{n,s,t}\le \bar g_{n,s,t}G_{n,s}$? The model could produce electricity without paying for the capacity needed to generate it.

\newpage

## 5. Motivate the solution: compare complete ways to meet demand

- The cheapest generator in isolation need not be part of the cheapest
electricity system. 
- The relevant comparison includes the infrastructure needed to deliver electricity at the time and place of demand.

### A two-country thought experiment

Suppose each country needs 1 MW in each of two one-hour periods. Existing
wind capacity could produce the following outputs:

| Available wind output (MW) | Morning | Evening |
| :------------------------- | ------: | ---: |
| Country A                  |       1 |    1 |
| Country B                  |       1 |    1 |
| Combined demand            |       2 |    2 |

- System is right-sized for each country. Stable production and demand

Suppose instead that the two countries have complementary production patterns (renewables):

| Available wind output (MW) | Morning | Evening |
|:--|--:|--:|
| Country A | 2 | 0 |
| Country B | 0 | 2 |
| Combined demand | 2 | 2 |

- A produces 2 MW in the morning and 0 MW in the evening.
- B produces 0 MW in the morning and 2 MW in the evening.
- With a 1 MW connection, A exports 1 MW in hour 1 and B exports 1 MW in hour 2. 
- No storage or backup is needed in these two hours. 
- Without the connection, each country must bridge its own shortfall using storage or backup. 
- With lossy storage, the given wind energy alone is insufficient over a cyclic
two-hour horizon: 

**Change the example:** Put both wind surpluses in hour 1. A connection
cannot move electricity to hour 2. Storage or dispatchable generation is
now necessary even with unrestricted trade. The value of transmission
depends on the spatial correlation of generation, not just its annual sum.

| Available wind output (MW) | Morning | Evening |
|:--|--:|--:|
| Country A | 2 | 2 |
| Country B | 0 | 0 |
| Combined demand | 2 | 2 |


### What the optimizer trades off

- **More local renewables:** reduce dependence on imports, but may require
  storage, backup, or curtailment during high-output hours.
- **More storage:** shift supply across hours, paying capacity costs and
  losing some energy in each cycle.
- **More transmission:** share geographically diverse output and access
  productive sites, paying construction costs and using the line-volume cap.
- **More gas generation:** cover difficult hours, but use the emissions budget.

The optimizer evaluates these options jointly across all hours. Building
capacity that is attractive in one hour can be unattractive over the year;
rare shortfalls can nevertheless justify capacity with low utilization.

The paper solves this sparse LP with PyPSA and Gurobi's barrier algorithm.
Students do not need to reproduce the large solve to understand its logic:
the primal solution identifies a feasible least-cost plan, and the dual
values measure the costs of tightening its constraints.

### Predict before seeing the results

If the transmission cap is relaxed while demand, weather, costs, and the
carbon budget remain fixed, minimum total cost cannot rise: the previous
solution remains feasible. Predict more use of productive wind locations
and less reliance on local storage. Individual countries or technologies
need not change monotonically even though minimum total cost does.

\newpage

## 6. Follow one sensitivity analysis: relax the transmission cap

The experiment changes only $\mathrm{CAP}_{LV}$ in $\sum_\ell l_\ell F_\ell\le \mathrm{CAP}_{LV}$.
For each value, **re-optimize all investment and operating decisions**.
Keep weather, demand, technology costs, resource bounds, and the CO$_2$
budget fixed. Do not hold the generation portfolio at its initial optimum.

| Scenario | Line volume (TWkm) | Average system cost (EUR/MWh) |
|:--|--:|--:|
| No international transmission | 0.00 | 84.1 |
| Existing-grid-equivalent volume | 31.25 | 75.7 |
| Compromise grid | 125.00 | 67.5 |
| Economically optimal grid | 285.70 | 64.8 |

Source: paper, Table 2. The existing-grid-equivalent scenario optimally
allocates the historical aggregate volume; individual line capacities need
not match the actual grid. TWkm denotes a million MWkm.

```{r}
#| label: fig-transmission-sensitivity
#| fig-cap: "Four published scenarios from Table 2. Dashed segments connect observations as visual guides; they do not reconstruct the full model's sensitivity curve or its LP breakpoints."
#| fig-width: 6.3
#| fig-height: 2.5
#| fig-pos: H
#| dev: pdf
par(mar = c(4, 4.4, 0.4, 0.5), mgp = c(2.6, 0.8, 0))
plot(volume, cost, type = "o", pch = 19, lty = 2,
     col = "#246080", lwd = 1.5, xlim = c(0, 305), ylim = c(62, 88),
     xlab = "Allowed transmission volume (TWkm)",
     ylab = "System cost (EUR/MWh)", las = 1, cex.lab = 0.9)
text(volume, cost, labels = c("Zero", "Existing volume", "Compromise", "Optimum"),
     pos = c(4, 4, 3, 3), cex = 0.8, offset = 0.5)
```

### Quantify the compromise

Relative to the zero-transmission benchmark, the maximum reported reduction
is $84.1-64.8=19.3$ EUR/MWh. The compromise captures $84.1-67.5=16.6$:

$$
\text{Savings share}=\frac{16.6}{19.3}\approx86\%,\qquad
\text{Volume share}=\frac{125}{285.7}\approx44\%.
$$

The paper reports approximately 85% of savings; 86% follows from its rounded
table entries. Against the existing-volume benchmark instead, the savings
share is $(75.7-67.5)/(75.7-64.8)\approx75\%$. State the benchmark.

**Interpretation:** Much of the gain is available before the economic
optimum. Restricting transmission sharply favors solar and local storage;
relaxing the cap permits more wind and geographic balancing. These are
endogenous responses to the cap, not imposed changes in technology shares.

\newpage

## 7. Use the shadow price to explain the sensitivity result

For the sensitivity derivation, let $C^*(\mathrm{CAP}_{LV})$ be minimized **annual total cost** (teaching notation). Collect the transmission-cap
term in the Lagrangian, with nonnegative multiplier $\mu_{LV}$:

$$
\mathcal L=C(x)+\mu_{LV}\left(\sum_\ell l_\ell F_\ell-\mathrm{CAP}_{LV}\right)
+\text{terms for the other constraints}.
$$

Where the value function is differentiable, the envelope result gives

$$
\frac{dC^*(\mathrm{CAP}_{LV})}{d\mathrm{CAP}_{LV}}=-\mu_{LV},\qquad
\Delta C^*\approx-\mu_{LV}\,\Delta\mathrm{CAP}_{LV}.
$$

The multiplier measures annual euros saved by permitting one additional
MWkm. Construction costs are already in the objective: this is a **net
system saving**, not avoided storage cost alone.

### Why the marginal value falls

First, more allowed line volume expands the feasible set, so
$C^*(\mathrm{CAP}_{LV})$ is nonincreasing. Second, an LP's value function is piecewise
linear and convex in this right-hand-side parameter. Its slope becomes
less negative as $\mathrm{CAP}_{LV}$ increases; correspondingly, the marginal saving
$\mu_{LV}$ weakly declines. At a corner there may be multiple optimal dual
values, so use the appropriate one-sided slope.

The paper's many-scenario curve can look smooth even though the underlying
LP value function is piecewise linear. **A changing slope across scenarios
does not make the optimization problem nonlinear.**

Complementary slackness gives

$$
\mu_{LV}\left(\mathrm{CAP}_{LV}-\sum_\ell l_\ell F_\ell\right)=0.
$$

Once the cap is above the unconstrained cost-optimal line volume, it is
slack and $\mu_{LV}=0$. Permitting still more capacity has no value because
the model can already build its preferred network. A binding constraint
does not, by itself, guarantee a strictly positive multiplier.

### Finite changes require more than one local shadow price

For a large relaxation, re-solve as the active constraints change:

$$
C^*(\mathrm{CAP}_{LV}^{(0)})-C^*(\mathrm{CAP}_{LV}^{(1)})=\int_{\mathrm{CAP}_{LV}^{(0)}}^{\mathrm{CAP}_{LV}^{(1)}}\mu_{LV}(u)\,du.
$$

The Table 2 endpoints give finite changes in reported average costs, not
raw dual values. Neither a secant slope between those scenarios nor its
units should be confused with the annual EUR/MWkm multiplier above.
To check a solver's dual numerically, compare annual objective values from
nearby re-solves using a consistent MWkm scale and sign convention.

**Connection to the previous application:** A right-hand-side change alters
the feasible set and the optimal allocation. Its value depends on what is
allowed to adjust. Here a simple upper-bound relaxation always preserves
the old feasible choices; it carries no obligation to build the extra grid.

\newpage

## 8. Interpret the result and check understanding

### What can be concluded?

Under the paper's assumptions, international cooperation reduces the cost
of meeting a stringent emissions target. It allows countries to share
resources and manage variability across space. The benefits diminish, so a
grid smaller than the least-cost optimum can retain most of the savings.

The line-volume cap represents a restriction on development. It does not
estimate public opposition or environmental damages in euros. Consequently,
the compromise is an economically informative alternative, not a demonstrated
social optimum after accounting for all external costs.

The magnitude depends on the aggregation of each country to one node,
one year of known weather and demand, fixed storage durations, and the
technology menu and costs. The model omits internal network bottlenecks,
demand response, and flexibility from other energy sectors. These limitations
affect the numerical prescription while leaving the LP reasoning useful.

### Questions for discussion

1. If we require wind output to equal available output in every hour,
   which flexibility do we remove? How could that change feasibility?
2. Why can a carbon cap bind even when gas generation capacity is not fully
   used in most hours?
3. Why must generation and storage be re-optimized when $\mathrm{CAP}_{LV}$ changes?
4. If a more permissive transmission cap produces a higher minimized annual
   cost, what should we check before interpreting the result?
5. Can a country face a higher electricity price after integration even
   though Europe-wide total cost falls?

### Instructor checkpoints

1. Equality removes curtailment. Surplus output must be consumed, stored,
   or exported; otherwise the constraints can become infeasible.
2. Emissions restrict annual fossil energy, whereas capacity restricts
   instantaneous power. Gas capacity can serve rare shortage hours.
3. Holding other investments fixed values a different, narrower adjustment.
   The paper compares complete least-cost systems at each cap.
4. Check that only the upper bound changed, the previous plan remains
   feasible, both solves reached comparable optima, and the comparison uses
   the same annual objective rather than differently normalized measures.
5. Yes. An exporting region can face greater demand for its inexpensive
   resources. Aggregate savings do not establish how gains are distributed.

### Source and interpretation notes

- Model: paper Section 2, equations (1)--(10); assumptions: Section 3.
  Symbols follow the paper, including $G_{n,s}$ for generation and storage
  capacity, $soc_{n,s,t}$, $\bar g_{n,s,t}$, and $\mathrm{CAP}_{LV}$.
- Sensitivity: Section 4.1, Figure 2, and Table 2. The manuscript prose gives
  79.9 EUR/MWh for the compromise; these notes use Table 2's 67.5, which
  agrees with the stated savings relationship.
- Figure 8 expresses transmission value as an equivalent overnight capital
  cost. It is not directly the raw annual net multiplier derived here.
  The notes use the cap experiment and Table 2 to keep that distinction clear.
