---
title: "Model Lab: The Value of Land-Use Flexibility"
format:
  revealjs:
    theme: simple
    title-slide: false
    slide-number: true
    transition: fade
    center: false
    width: 1280
    height: 720
    margin: 0.04
    embed-resources: true
    toc: false
    number-sections: false
    include-in-header:
      text: |
        <style>
        .reveal { font-size: 32px; }
        .reveal h2 { font-size: 1.45em; margin-bottom: 0.6em; }
        .reveal h3 { font-size: 1.05em; margin-top: 0.5em; }
        .reveal p, .reveal li { line-height: 1.3; }
        .reveal .shared-assumptions { font-size: 0.85em; margin-top: 1.1em; }
        .reveal .valuation { text-align: center; font-size: 1.18em; margin: 0.8em 0; }
        .reveal .result { font-size: 0.9em; text-align: center; }
        .reveal .cell-output-display img { margin: 0 auto; }
        </style>
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: slide-calculations
#| include: false
# Reuse the model and audited calculations from the neighboring notes.
# Rendering this QMD requires Stewart_Notes.qmd, lpSolve, ggplot2, and knitr.
# The rendered HTML embeds its figure and presentation resources.
notes_source <- readLines("Stewart_Notes.qmd", warn = FALSE)
notes_chunk <- function(label) {
  start <- which(trimws(notes_source) == paste0("#| label: ", label))
  if (length(start) != 1L) stop("Expected one notes chunk named: ", label)
  closing <- which(seq_along(notes_source) > start &
                     trimws(notes_source) == "```")
  if (!length(closing)) stop("Missing closing fence for: ", label)
  notes_source[seq.int(start + 1L, min(closing) - 1L)]
}
for (label in c("setup-model", "lab-calculations")) {
  eval(parse(text = notes_chunk(label)), envir = knitr::knit_global())
}

```

## What is the value of unrestricted new acreage?

Add *t* acres of **group 2 land** and compare two versions of the same project.

:::: {.columns}
::: {.column width="50%"}
### Restricted expansion

On the additional acreage:

- Corn: at most 40%.
- Wheat: at most 30%.
- Pasture: at least 30%.
:::

::: {.column width="50%"}
### Unrestricted expansion

Any combination of corn, wheat, and pasture on the new land.

Total use cannot exceed the additional acreage available.
:::
::::

::: {.shared-assumptions}
**Existing land keeps its restrictions.** Both projects use the same yields,
production costs, and output requirements.
:::

::: {.notes}
This extends Stewart and Greenhalgh (1977). Only the treatment of newly added
group 2 acreage changes. Restricted shares are upper bounds for corn and wheat
and a lower bound for pasture, rather than three exact required shares.
Increasing the total-land bound alone while retaining the original aggregate
crop caps would not make the new land unrestricted. The notes implement
separate activity variables for the additional acres.
:::

## Policy relevance of land-use flexibility

A development program may require pasture or limit cropping on new land.
What production benefit does the program give up by attaching those rules?


$$
V(t) = C_R(t) - C_U(t) = B_U(t) - B_R(t)
$$


Here, **C** is minimized production cost and **B** is the saving from expansion.
**V** measures the value of removing the restrictions for a project of size *t*.

- Compare that production cost with the restrictions' conservation benefits.
- Evaluate the proposed project size: the value per acre changes with acreage.

::: {.notes}
R denotes restricted expansion and U denotes unrestricted expansion. Both
benefits use the same baseline production cost. Feasible-set inclusion implies
V(t) is nonnegative, but it need not rise at every project size. Environmental
benefits are outside this production model and must be evaluated separately.
The paper's coefficients are illustrative, so these dollars describe the
example rather than an estimate for a current real-world policy.
Source: Stewart_Notes.qmd, Model Lab Example.
:::

## Flexibility is worth \$`r money(ten$value)` for 10 added acres

```{r}
#| label: flexibility-gap-plot
#| fig-width: 10
#| fig-height: 4.2
#| out-width: "94%"
#| fig-align: center
#| fig-alt: "Restricted and unrestricted production-cost savings as group 2 acreage expands. The shaded gap is the value of flexibility. At 10 additional acres, the gap is 7.16 dollars."
#| dev: svg
# Recreate Figure 2 with the same computed values and larger slide labels.
ggplot(benefit_data, aes(t, benefit, color = scenario, linetype = scenario)) +
  geom_ribbon(data = benefit_band,
              aes(x = t, ymin = lower, ymax = upper),
              inherit.aes = FALSE, fill = "#DCE8EC", alpha = 0.8) +
  geom_line(linewidth = 1.15) +
  annotate("text", x = 19.8,
           y = baseline$cost - regime_cost(19.8, unrestricted_regimes) + 0.8,
           label = "Unrestricted", color = scenario_colors[["Unrestricted"]],
           hjust = 1, size = 5.6, fontface = "bold" ) +
  annotate("text", x = 19.8,
           y = baseline$cost - regime_cost(19.8, restricted_regimes) - 0.9,
           label = "Restricted", color = scenario_colors[["Restricted"]],
           hjust = 1, size = 5.6, fontface = "bold" ) +
  annotate("segment", x = 10, xend = 10,
           y = baseline$cost - ten$restricted,
           yend = baseline$cost - ten$unrestricted,
           linewidth = 0.8, color = "gray25") +
  annotate("text", x = 10.45,
           y = baseline$cost - (ten$restricted + ten$unrestricted) / 2,
           label = paste0("$", money(ten$value), " value of flexibility"),
           hjust = 0, size = 5) +
  scale_color_manual(values = scenario_colors) +
  scale_linetype_manual(values = scenario_lines) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Additional group 2 land (acres)",
       y = "Total cost saving ($)") +
  theme_minimal(base_size = 17) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "none",
        plot.margin = margin(5, 12, 0, 5))
```

::: {.result}
At 10 acres, total savings are **\$`r money(baseline$cost - ten$unrestricted)` unrestricted**
versus **\$`r money(baseline$cost - ten$restricted)` restricted**.
The **\$`r money(ten$value)` difference** is the value of flexibility.
:::

::: {.notes}
This reproduces Figure 2 from the notes. At any project size, the vertical gap
between the cost-saving curves is the total value of flexibility. At ten acres,
the difference is approximately $7.16, or $0.72 per added acre on average.
The unrestricted benefit levels off after about 8.38 acres. Restricted
benefits peak at 12.5 acres and then decline as the pasture minimum grows.
The shrinking and widening gap shows why the initial marginal value of
flexibility cannot be applied to every acre of a larger project. All monetary
values use the original model's accounting period, not capitalized land prices.
:::

## Land use restrictions carry an economic cost

- Unrestricted expansion allows landowners more flexibility in crop choice and production.
- These ideas could be applied to zoning, conservation easements, and other land-use regulations.

## Brainstorm 

- How might your modify the Stewart and Greenhalgh (1977) model?
- What aspects of the model interest you?
- Any implicit assumptions or limitations you would like to explore?
