---
title: "Helfand and House: A Code Walkthrough"
subtitle: "Classroom companion to HH_model.R"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    code-copy: true
    code-fold: false
    code-overflow: wrap
execute:
  eval: false
  echo: true
---

## How to use this tutorial

Open [HH_model.R](HH_model.R) in RStudio and run a few lines at a time. The
numbered sections below follow the script in order, with the same code and
object names. You can also use the **copy button** on each code block and
paste it into RStudio. Run the blocks from top to bottom in the same R
session: later blocks use objects created earlier. Code is displayed but
**never executed when these notes are rendered**; your R session supplies
the results we discuss in class.

The script needs R and the `nloptr` package. If the package is missing, run
this once in the R console:

```{r}
#| label: install-package
install.packages("nloptr")
```

The walkthrough is self-contained and can run from any working directory.
You do not need to source `model.R` or load a dataset first.

Our question is: **How does soil heterogeneity affect the cost of reducing
agricultural pollution?** Farmers choose nitrogen and irrigation on two
soils, Mocho and Pacheco. We first solve their unregulated choices, then
compare seven policies that target the same 20% reduction in average
leaching. Watch how each policy changes the incentives and the choices
available to farmers.

<!-- ::: {.callout-note title="The model we are using"}
This is the classroom reconstruction used in `HH_model.R`, not an exact
replication of the published tables. It uses a negative linear nitrogen
coefficient in Mocho's yield equation and omits the
$-1.11\sqrt{N}$ term in Pacheco's leaching equation. These are reconstruction
choices, not author-confirmed corrections. The [README](README.md) records
the source discrepancies. Our task here is to understand the supplied
model and code.
::: -->

## 1. Prices, areas, and biological responses

### Set prices and describe the two soils

`library(nloptr)` loads the optimization tools. The assignment operator
`<-` stores a value under a name; `c()` combines values into a vector.
`area_share` is a named vector giving the fraction of total land on each
soil. The paper assumes an equal distribution of lettuce fields across the
two soil types (p. 1027), so each share is 0.5.

In these notes, we write those shares as $a_s$, where $s$ identifies the
soil: $a_M=a_P=0.5$ and $a_M+a_P=1$. **The symbol $a_s$ is our notation for
the paper's equal-area assumption.** These shares are fixed land weights,
not estimated biological coefficients or choices that farmers optimize.

| Quantity | Units in the script |
|---|---|
| Nitrogen, `N` | kg per hectare |
| Irrigation depth, `W` | mm |
| Yield | tons of lettuce biomass per hectare |
| Leaching | kg per hectare |
| `lettuce_price` | dollars per ton, net of harvest cost |
| `nitrogen_price` | dollars per kg |
| `water_price` | dollars per mm-ha |
| Quasi-rent | dollars per hectare |

```{r}
#| label: hh-01-1
library(nloptr)

lettuce_price <- 1350                   # dollars per ton, net of harvest cost
nitrogen_price <- 0.70                  # dollars per kg
water_price <- 0.23                     # dollars per mm-ha
area_share <- c(Mocho = 0.5, Pacheco = 0.5)
```

### Define the biological response functions

`function(N, W)` creates a reusable calculation. Defining a function does
not evaluate it; calling it with input values does. R returns the last
expression in the function body.

The first two functions predict yield; the next two predict leaching.
Their coefficients differ by soil. The `N*W` interaction means that the
marginal effect of nitrogen depends on water use, and vice versa.
`sqrt(N)` and `sqrt(W)` let marginal responses change with input levels.
A negative linear coefficient alone does not tell us the overall marginal
effect: we must include the interaction and square-root terms.

```{r}
#| label: hh-01-2
# N is kg/ha, W is irrigation depth in mm.
# Yield is tons of dry lettuce biomass/ha; leaching is kg/ha.
yield_mocho <- function(N, W) {
  0.9117 - 0.00363*N - 0.0029*W + 0.00000193*N*W +
    0.0506*sqrt(N) + 0.156*sqrt(W)
}

yield_pacheco <- function(N, W) {
  2.296 - 0.00195*N - 0.00158*W + 0.00000153*N*W +
    0.0249*sqrt(N) + 0.0732*sqrt(W)
}

leaching_mocho <- function(N, W) {
  -26.06 - 0.152*N + 0.158*W + 0.000363*N*W
}

leaching_pacheco <- function(N, W) {
  43.82 - 0.17*N + 0.313*W + 0.000466*N*W - 7.28*sqrt(W)
}
```

### Evaluate an allocation before optimizing

The first two calls evaluate Mocho at a particular input combination.
They do not search for better inputs. `starting_inputs` then collects all
four choices in the order used throughout the script:

$$x=(N_M,W_M,N_P,W_P).$$

`farm_results()` rearranges that vector into a two-row table. R indexes
from 1, so `x[1]` and `x[3]` select the two nitrogen choices. The function
calculates quasi-rent on each soil as

$$q_s(N_s,W_s)=p_Y Y_s(N_s,W_s)-c_NN_s-c_WW_s.$$

Quasi-rent subtracts nitrogen and irrigation costs; it does not deduct all
farm costs. `data.frame()` collects the inputs and outcomes into columns.

```{r}
#| label: hh-01-3
# Try one point before optimizing. These are the published Mocho inputs.
yield_mocho(N = 85, W = 723)
leaching_mocho(N = 85, W = 723)

# x always contains FOUR choices in this order: N_M, W_M, N_P, W_P.
starting_inputs <- c(N_M = 85, W_M = 723, N_P = 52, W_P = 478)

# Translate four input choices into a two-row table of economic outcomes.
farm_results <- function(x) {
  N <- c(x[1], x[3])
  W <- c(x[2], x[4])
  yield <- c(yield_mocho(N[1], W[1]), yield_pacheco(N[2], W[2]))
  leaching <- c(leaching_mocho(N[1], W[1]), leaching_pacheco(N[2], W[2]))
  rent <- lettuce_price*yield - nitrogen_price*N - water_price*W
  #output
  data.frame(soil = c("Mocho", "Pacheco"), N, W, yield, leaching, rent,
             row.names = NULL)
}

farm_results(starting_inputs)
```

### Convert soil outcomes to landscape averages

`farm_results(x)$rent` extracts the table's rent column. Multiplication by
`area_share` weights each soil's outcome, and `sum()` adds the contributions:

$$Q(x)=\sum_s a_s q_s(x_s),\qquad L(x)=\sum_s a_s L_s(x_s).$$

These are averages per hectare of the whole landscape, not totals for an
unspecified farm acreage. With equal areas they are simple averages:

$$L(x)=0.5L_M(x_M)+0.5L_P(x_P).$$

For example, if Mocho leaches 100 kg/ha and Pacheco leaches 40 kg/ha, the
landscape average is $0.5(100)+0.5(40)=70$ kg/ha. The same land shares weight
quasi-rent. Their purpose is to give each soil's per-hectare outcome its
appropriate contribution to the landscape average.

```{r}
#| label: hh-01-4
# Quasi-rent excludes costs other than fertilizer and irrigation.
# Area shares convert soil-specific outcomes to landscape averages per ha.
mean_rent <- function(x) sum(area_share * farm_results(x)$rent)
mean_leaching <- function(x) sum(area_share * farm_results(x)$leaching)

mean_rent(starting_inputs)
mean_leaching(starting_inputs)
```

**Before moving on:** Identify the four entries in `starting_inputs`, the two rows in `farm_results()`, and the single value returned by `mean_leaching()`. These are three views of the same allocation.

## 2. Optimize the unregulated baseline

We now choose the four inputs to maximize $Q(x)$. With no shared pollution
constraint, maximizing this weighted sum also solves each soil's private
problem independently.

### Set numerical bounds and scaling

`rep(1e-6, 4)` repeats a small positive lower bound four times. The floor
avoids the undefined derivative of a square root at zero. `input_upper`
limits the search to the reconstruction's numerical domain; these bounds
are distinct from the policy ceilings introduced later.

The solver works with scaled choices $z$, where
`x = z * input_scale` (element-by-element multiplication).
Dividing nitrogen by 100 and water by 1,000 gives variables of more similar
size. This changes their numerical representation, not their physical units
when we report results. The fitted response functions should not be
extrapolated arbitrarily beyond this domain.

```{r}
#| label: hh-02-1
# SLSQP is sequential quadratic programming: approximate the objective
# locally by a quadratic and constraints by linear functions, solve that
# subproblem, update the input choices, and repeat.

# Keep the original model's numerical domain and scaling.
# These are computational choices, not additional policies from the paper.
input_lower <- rep(1e-6, 4)
input_upper <- c(150, 1000, 100, 700)
input_scale <- c(100, 1000, 100, 1000)
```

### Supply the objective and its derivatives

`nloptr()` minimizes an objective, so we minimize **negative** average
quasi-rent. Dividing it by 1,000 further adjusts numerical scale without
changing its maximizer.

`rent_gradient(x)` returns the four derivatives of average quasi-rent with
respect to the physical inputs. For example,

$$\frac{\partial Q}{\partial N_M}
=a_M\left[p_Y\left(-0.00363+0.00000193W_M+
\frac{0.0506}{2\sqrt{N_M}}\right)-c_N\right].$$

Each component is the area-weighted value of marginal yield minus the
input price. `rep(area_share, each = 2)` gives weights in input order:
$(a_M,a_M,a_P,a_P)$. By contrast, `rep(c(nitrogen_price, water_price), 2)`
repeats the whole price pair.

```{r}
#| label: hh-02-2
# SLSQP minimizes. Negate rent to maximize it.
# The optimizer chooses z; x = z*input_scale converts back to physical units.
# SLSQP uses first derivatives. These are the four marginal quasi-rents.
# Writing them explicitly keeps square-root derivatives inside their domain.
rent_gradient <- function(x) {
  N_M <- x[1]; W_M <- x[2]; N_P <- x[3]; W_P <- x[4]
  yield_slopes <- c(
    -0.00363 + 0.00000193*W_M + 0.0506/(2*sqrt(N_M)),
    -0.0029 + 0.00000193*N_M + 0.156/(2*sqrt(W_M)),
    -0.00195 + 0.00000153*W_P + 0.0249/(2*sqrt(N_P)),
    -0.00158 + 0.00000153*N_P + 0.0732/(2*sqrt(W_P))
  )
  rep(area_share, each = 2) *
    (lettuce_price*yield_slopes - rep(c(nitrogen_price, water_price), 2))
}
baseline_objective <- function(z) -mean_rent(z * input_scale) / 1000
```

### Ask SLSQP to find the baseline

SLSQP (sequential quadratic programming) repeatedly builds and solves a
local approximation to the optimization problem, then updates the choices.

| Argument | Meaning here |
|---|---|
| `x0` | Initial guess in scaled units |
| `eval_f` | Function calculating the objective |
| `eval_grad_f` | Function calculating the objective's gradient |
| `lb`, `ub` | Lower and upper bounds in scaled units |
| `algorithm` | Selects SLSQP |
| `xtol_rel`, `ftol_abs` | Stopping tolerances for changes in choices and objective |
| `maxeval` | Limit on objective evaluations |

The gradient callback includes a minus sign, division by 1,000, and
multiplication by `input_scale`. The last factor follows from the chain
rule because the solver changes $z$, while `rent_gradient()` differentiates
with respect to $x$. `function(z)` defines that callback directly inside
the solver call.

The returned object contains more than the solution. When running the
script, inspect `baseline_fit$status` and `baseline_fit$message` if you
need to diagnose convergence. A returned vector alone does not establish
that an optimizer succeeded.

```{r}
#| label: hh-02-3
baseline_fit <- nloptr(
  x0 = starting_inputs / input_scale,
  eval_f = baseline_objective,
  eval_grad_f = function(z) -rent_gradient(z * input_scale) * input_scale / 1000,
  lb = input_lower / input_scale,
  ub = input_upper / input_scale,
  opts = list(algorithm = "NLOPT_LD_SLSQP", xtol_rel = 1e-10,
              ftol_abs = 1e-12, maxeval = 2000)
)
```

## 3. Look at the baseline

### Recover physical inputs and set the target

`$solution` extracts the solver's chosen vector. Multiplying by
`input_scale` converts it back to kg/ha and mm. Printing
`farm_results(baseline_inputs)` shows whether the two soils choose different
inputs even though they face the same prices.

Save baseline quasi-rent and leaching before introducing any policy.
Every policy uses the common target

$$\bar L=0.8L(x^0),$$

where $x^0$ is the computed baseline. A 20% reduction in **average leaching**
does not require a 20% reduction in every input or on every soil.

```{r}
#| label: hh-03-1
baseline_inputs <- baseline_fit$solution * input_scale
baseline_inputs
farm_results(baseline_inputs)
baseline_rent <- mean_rent(baseline_inputs)
baseline_leaching <- mean_leaching(baseline_inputs)
baseline_rent
baseline_leaching

# Every policy below targets the same 20% reduction.
leaching_target <- 0.8 * baseline_leaching
leaching_target
```

### Keep the policy accounting consistent

`policy_result()` builds one row for the final comparison. Its default
`taxes = rep(0, 4)` means zero taxes unless a vector is supplied. The tax
vector always follows $(N_M,W_M,N_P,W_P)$.

For each soil, tax payments equal each per-unit rate times its input use.
Area weights then turn those payments into dollars per landscape hectare.
The accounting is

$$\text{efficiency cost}=Q(x^0)-Q(x),$$
$$\text{farmer net rent}=Q(x)-\text{tax payments}.$$

Tax payments reduce farmers' receipts but are transfers in this model's
efficiency accounting. We therefore do not add them to efficiency cost.
Administrative costs and environmental damages are not included. The
baseline row should have zero efficiency cost and zero tax payments.

```{r}
#| label: hh-03-2
# One small reporting function lets us compare all the policies consistently.
# Taxes are transfers; efficiency cost is lost PRE-TAX quasi-rent.
policy_result <- function(policy, x, taxes = rep(0, 4)) {
  tax_by_soil <- c(sum(taxes[1:2] * x[1:2]), sum(taxes[3:4] * x[3:4]))
  transfers <- sum(area_share * tax_by_soil)
  data.frame(policy,
             efficiency_cost = baseline_rent - mean_rent(x),
             leaching = mean_leaching(x),
             tax_payments = transfers,
             farmer_net_rent = mean_rent(x) - transfers)
}

baseline_result <- policy_result("Baseline", baseline_inputs)
baseline_result
```

## 4. Efficient allocation and differentiated taxes

### Impose the environmental target directly

For the efficient benchmark, a planner chooses all four inputs:

$$\max_x Q(x)\quad\text{subject to}\quad L(x)\leq\bar L.$$

The objective is unchanged. The new ingredient is `pollution_constraint()`.
`nloptr()` expects inequalities in the form $g(z)\leq0$, so the function
returns leaching **minus** the target, divided by 100 for scaling.
A positive value means excess pollution; zero means the cap binds.

`eval_g_ineq` supplies this constraint, and `eval_jac_g_ineq` supplies its
derivatives. `nl.grad()` approximates derivatives numerically by evaluating
nearby points. The initial guess is the unregulated baseline, which exceeds
the pollution target; the constrained solver must move to a feasible
allocation. Compare the resulting average leaching with `leaching_target`.

```{r}
#| label: hh-04-1
# The planner chooses all four inputs. nloptr expects g(z) <= 0.
efficient_objective <- function(z) -mean_rent(z * input_scale) / 1000
pollution_constraint <- function(z) {
  (mean_leaching(z * input_scale) - leaching_target) / 100
}

efficient_fit <- nloptr(
  x0 = baseline_inputs / input_scale,
  eval_f = efficient_objective,
  eval_grad_f = function(z) -rent_gradient(z * input_scale) * input_scale / 1000,
  eval_g_ineq = pollution_constraint,
  eval_jac_g_ineq = function(z) nl.grad(z, pollution_constraint),
  lb = input_lower / input_scale,
  ub = input_upper / input_scale,
  opts = list(algorithm = "NLOPT_LD_SLSQP", xtol_rel = 1e-10, maxeval = 2000)
)

efficient_inputs <- efficient_fit$solution * input_scale
farm_results(efficient_inputs)
mean_leaching(efficient_inputs)
```

### Translate the efficient allocation into input taxes

Using the Lagrangian $Q(x)+\lambda[\bar L-L(x)]$, an interior input choice
satisfies

$$\frac{\partial Q}{\partial x_i}
=\lambda\frac{\partial L}{\partial x_i}.$$

The code recovers $\lambda$ from the Mocho nitrogen condition by taking a
ratio of derivatives. This calculation assumes that input is interior,
the pollution constraint binds, and its marginal leaching is nonzero; an
active input bound would introduce another term into the condition.

`shadow_value` is the marginal increase in maximized quasi-rent from
relaxing the leaching target by one unit: dollars per kg of allowed
leaching. Both derivatives in the ratio carry the same area weight, so it
cancels. These derivatives use physical inputs and unscaled outcomes;
there is no need to undo the solver's objective or constraint scaling.

A supporting per-unit input tax on soil $s$ is

$$t_{Ns}=\lambda\frac{\partial L_s}{\partial N_s},\qquad
 t_{Ws}=\lambda\frac{\partial L_s}{\partial W_s},$$

evaluated at the efficient allocation. `leaching_slopes` contains derivatives
of **average** leaching, so dividing by `rep(area_share, each = 2)` removes
the area weights. Nitrogen rates are dollars/kg; water rates are dollars/mm-ha.
The resulting four rates can differ across soils because marginal leaching
differs. Section 5 checks whether farmers facing these rates choose the
planner's allocation.

```{r}
#| label: hh-04-2
# Recover lambda from the nitrogen FOC on Mocho: q_N = lambda * L_N.
# Reuse the rent derivatives; nl.grad estimates the leaching derivatives.
rent_slopes <- rent_gradient(efficient_inputs)
leaching_slopes <- nl.grad(efficient_inputs, mean_leaching)
shadow_value <- rent_slopes[1] / leaching_slopes[1]
shadow_value                          # dollars per kg of allowed leaching

# Differentiated taxes: lambda times each soil's marginal leaching.
# Remove area weights from the slopes of LANDSCAPE-average leaching.
differentiated_taxes <- shadow_value * leaching_slopes /
  rep(area_share, each = 2)
names(differentiated_taxes) <- names(starting_inputs)
differentiated_taxes

efficient_result <- policy_result("Differentiated taxes", efficient_inputs,
                                  differentiated_taxes)
efficient_result
```

## 5. Farmer responses to a tax or input ceiling

`choose_inputs()` is the producer-response function used by all remaining
policies. Give it four tax rates, four ceilings, or both; it returns four
chosen inputs in physical units. By default it applies no taxes and retains
the numerical upper bounds.

Inside the function, farmers maximize **after-tax** quasi-rent:

$$\sum_s a_s[q_s(N_s,W_s)-t_{Ns}N_s-t_{Ws}W_s].$$

`after_tax_objective()` and `after_tax_gradient()` can use the `taxes`
argument supplied to the surrounding function. The gradient subtracts the
area-weighted tax rates from the marginal pre-tax returns. Because soils
have separate technologies and no shared constraint in this step, one
optimization solves both farmers' choices.

Ceilings enter through `ub`. `pmin(baseline_inputs, ceilings)` takes the
smaller value at each position to keep the initial guess within those
ceilings. A ceiling restricts an input to be **at most** that amount; it
does not force the farmer to use it.

The last expression returns `fit$solution * input_scale`. Compare the two
vectors printed at the end: `choose_inputs(taxes = differentiated_taxes)`
and `efficient_inputs` should be close if the supporting taxes decentralize
the efficient allocation and both solves converge.

```{r}
#| label: hh-05-1
# Reuse the baseline optimization, now allowing taxes and policy ceilings.
# Maximizing aggregate farmer receipts here also solves each farmer's problem:
# farmers have separate technologies and no shared constraint in this step.
choose_inputs <- function(taxes = rep(0, 4), ceilings = input_upper) {
  after_tax_objective <- function(z) {
    x <- z * input_scale
    tax_by_soil <- c(sum(taxes[1:2]*x[1:2]), sum(taxes[3:4]*x[3:4]))
    -(mean_rent(x) - sum(area_share * tax_by_soil)) / 1000
  }
  after_tax_gradient <- function(z) {
    -(rent_gradient(z * input_scale) - rep(area_share, each = 2) * taxes) *
      input_scale / 1000
  }
  fit <- nloptr(
    x0 = pmin(baseline_inputs, ceilings) / input_scale,
    eval_f = after_tax_objective,
    eval_grad_f = after_tax_gradient,
    lb = input_lower / input_scale,
    ub = ceilings / input_scale,
    opts = list(algorithm = "NLOPT_LD_SLSQP", xtol_rel = 1e-10,
                ftol_abs = 1e-12, maxeval = 2000)
  )
  fit$solution * input_scale
}

# Example: farmers facing the differentiated taxes choose the efficient inputs.
choose_inputs(taxes = differentiated_taxes)
efficient_inputs
```

## 6. Uniform taxes on both inputs

### Separate the regulator's choice from the farmers' choices

Now there are only two tax rates: one for nitrogen and one for water,
shared by both soils. The regulator cannot set four independent rates.
`rep(z * tax_scale, 2)` expands the two scaled rates into
$(t_N,t_W,t_N,t_W)$ before calling `choose_inputs()`.

This is a nested problem. For each candidate tax pair, farmers choose
inputs; the regulator evaluates the resulting quasi-rent and leaching:

$$\max_{t_N,t_W\geq0}Q(x(t_N,t_W))
\quad\text{subject to}\quad L(x(t_N,t_W))\leq\bar L.$$

The inner farmer problem subtracts taxes. The outer regulator objective
uses **pre-tax** quasi-rent, consistent with treating tax revenue as a
transfer. Here `z` represents two scaled tax rates, rather than the four
scaled inputs used in earlier solver calls.

```{r}
#| label: hh-06-1
# Regulator chooses TWO rates, repeated across soils: (t_N, t_W, t_N, t_W).
# Each trial pair requires farmers to re-optimize all four input choices.
tax_scale <- c(1, 0.2)
uniform_response <- function(z) {
  choose_inputs(taxes = rep(z * tax_scale, 2))
}
uniform_objective <- function(z) -mean_rent(uniform_response(z)) / 1000
uniform_constraint <- function(z) {
  (mean_leaching(uniform_response(z)) - leaching_target) / 100
}
```

### Find the tax pair and inspect the induced allocation

Each call to `uniform_objective()` or `uniform_constraint()` triggers a
farmer optimization. Numerical derivatives therefore measure how outcomes
change after farmers re-optimize in response to a changed tax rate.
`heps = 1e-4` sets the finite-difference step used by `nl.grad()`; the outer
solver also specifies a constraint tolerance.

`lb = c(0, 0)` restricts both tax rates to be nonnegative. After solving,
`uniform_fit$solution * tax_scale` recovers rates in actual price units.
The code then solves the farmers' problem at those rates and records the
policy outcome. This nested calculation may take more time than a single
farmer solve.

**Pause and discuss:** Why can uniform tax rates lead to different input
choices on the two soils? What flexibility does the regulator lose relative
to differentiated taxes?

```{r}
#| label: hh-06-2
# SLSQP again; nl.grad estimates how the regulator's objective and constraint
# change when the tax rates change and farmers re-optimize.
uniform_fit <- nloptr(
  x0 = c(0.06, 0.9),
  eval_f = uniform_objective,
  eval_grad_f = function(z) nl.grad(z, uniform_objective, heps = 1e-4),
  eval_g_ineq = uniform_constraint,
  eval_jac_g_ineq = function(z) nl.grad(z, uniform_constraint, heps = 1e-4),
  lb = c(0, 0),
  opts = list(algorithm = "NLOPT_LD_SLSQP", xtol_rel = 1e-9,
              maxeval = 5000, tol_constraints_ineq = 1e-9)
)

uniform_rates <- uniform_fit$solution * tax_scale
names(uniform_rates) <- c("nitrogen", "water")
uniform_rates
uniform_inputs <- choose_inputs(taxes = rep(uniform_rates, 2))
farm_results(uniform_inputs)
uniform_result <- policy_result("Uniform taxes on both inputs", uniform_inputs,
                                rep(uniform_rates, 2))
uniform_result
```

## 7. Uniform nitrogen-only tax

With only one tax rate to determine, the script uses root finding. Define

$$h(t)=L(x(t,0))-\bar L.$$

`nitrogen_tax_gap(t)` calculates this gap after farmers choose both inputs.
A positive gap means the policy misses the target; a negative gap means it
reduces leaching beyond the target. `uniroot()` searches for a rate where
the gap is zero, and `$root` extracts that rate.

The interval `c(0, 15)` brackets the root for this specification: the endpoint
gaps must have opposite signs (or an endpoint must already be a root).
It is a search interval, not a policy restriction. If prices, coefficients,
or the target change, check the endpoint gaps before reusing the interval.
A bracketed root search finds a target-achieving rate; it is not a general
proof that responses are monotone or the root is unique.

The vector `c(t, 0, t, 0)` taxes nitrogen on both soils. **Water remains a
choice**, so a nitrogen-only tax can also change irrigation. The final
assignment saves a comparison row; assignments do not automatically print
in R, so type `nitrogen_tax_result` to inspect it immediately.

```{r}
#| label: hh-07-1
# With one instrument, find the rate that makes leaching equal the target.
# The intervals in these examples bracket the target for THIS specification.
# They are search intervals, not economic limits on the instruments.
nitrogen_tax_gap <- function(t) {
  x <- choose_inputs(taxes = c(t, 0, t, 0))
  mean_leaching(x) - leaching_target
}
nitrogen_rate <- uniroot(nitrogen_tax_gap, interval = c(0, 15), tol = 1e-8)$root
nitrogen_rate
nitrogen_tax_inputs <- choose_inputs(taxes = c(nitrogen_rate, 0, nitrogen_rate, 0))
farm_results(nitrogen_tax_inputs)
nitrogen_tax_result <- policy_result("Uniform nitrogen-only tax", nitrogen_tax_inputs,
                                     c(nitrogen_rate, 0, nitrogen_rate, 0))
```

## 8. Uniform water-only tax

This repeats the root-finding logic with `c(0, t, 0, t)`. The nitrogen tax
is zero, but farmers still re-optimize nitrogen along with water. The
search interval is now `c(0, 1)`, in dollars per mm-ha.

Compare `farm_results(water_tax_inputs)` with the nitrogen-tax allocation.
Look at both columns of inputs and both soils' leaching. The cost of an
instrument depends on how it changes the full production decision, not
just the quantity it directly taxes. Both policies seek the same average
leaching target, making their efficiency costs comparable.

```{r}
#| label: hh-08-1
water_tax_gap <- function(t) {
  x <- choose_inputs(taxes = c(0, t, 0, t))
  mean_leaching(x) - leaching_target
}
water_rate <- uniroot(water_tax_gap, interval = c(0, 1), tol = 1e-8)$root
water_rate
water_tax_inputs <- choose_inputs(taxes = c(0, water_rate, 0, water_rate))
farm_results(water_tax_inputs)
water_tax_result <- policy_result("Uniform water-only tax", water_tax_inputs,
                                  c(0, water_rate, 0, water_rate))
```

## 9. Common nitrogen ceiling

We now regulate a quantity instead of raising an input price.
`c(cap, 1000, cap, 700)` applies the same nitrogen ceiling to both soils
and leaves their water bounds at the original numerical values.

`nitrogen_cap_gap()` solves the farmer problem under each proposed cap.
`uniroot()` adjusts that cap until average leaching equals the target.
The interval `c(5, 85)` is measured in kg of nitrogen per hectare.

A common ceiling means $N_M\leq\text{cap}$ and
$N_P\leq\text{cap}$, not that nitrogen use must be identical. Compare the
chosen quantities to `nitrogen_cap` to see which constraints bind.
`policy_result()` receives no tax vector, so tax payments default to zero.
Farmers can still lose quasi-rent because the ceiling restricts their choices.

```{r}
#| label: hh-09-1
nitrogen_cap_gap <- function(cap) {
  x <- choose_inputs(ceilings = c(cap, 1000, cap, 700))
  mean_leaching(x) - leaching_target
}
nitrogen_cap <- uniroot(nitrogen_cap_gap, interval = c(5, 85), tol = 1e-8)$root
nitrogen_cap
nitrogen_cap_inputs <- choose_inputs(ceilings = c(nitrogen_cap, 1000, nitrogen_cap, 700))
farm_results(nitrogen_cap_inputs)
nitrogen_cap_result <- policy_result("Nitrogen cap", nitrogen_cap_inputs)
```

## 10. Common water ceiling

The policy ceiling applies to water on both soils. The `pmin()` operation
compares the proposed ceilings with `input_upper` element by element. This
keeps a proposed water cap above 700 from inadvertently expanding Pacheco's
original numerical domain.

The root search chooses a cap in mm that achieves the common leaching
target. After running it, compare the two water choices with `water_cap`.
As the script notes, Pacheco can use less than the ceiling: a common maximum
does not require equal irrigation. Nitrogen remains adjustable.

**Pause and discuss:** How could a water tax affect a farmer whose water use
would be below a common water ceiling? Why might that change the allocation
across soils even when aggregate leaching is the same?

```{r}
#| label: hh-10-1
water_cap_gap <- function(cap) {
  x <- choose_inputs(ceilings = pmin(input_upper, c(150, cap, 100, cap)))
  mean_leaching(x) - leaching_target
}
water_cap <- uniroot(water_cap_gap, interval = c(500, 723), tol = 1e-8)$root
water_cap
water_cap_inputs <- choose_inputs(ceilings = pmin(input_upper, c(150, water_cap, 100, water_cap)))
farm_results(water_cap_inputs)           # Pacheco can use less than the ceiling
water_cap_result <- policy_result("Water cap", water_cap_inputs)
```

## 11. Water and nitrogen caps as percentages of baseline

This policy uses a common reduction fraction $r$ to set four ceilings:

$$x_i\leq(1-r)x_i^0.$$

The percentage is common, but the numerical ceilings differ because they
are based on each soil's own baseline choices. For example, $r = 0.1$
would set each ceiling at 90% of that input's baseline quantity.

`both_caps_gap()` lets farmers optimize below those ceilings, then measures
the leaching gap. `uniroot()` searches between zero and a 30% rollback.
The resulting `reduction_fraction` need not equal 0.2: a percentage reduction
in inputs is different from a percentage reduction in leaching.

`both_ceilings` stores the limits and `both_caps_inputs` stores actual
choices. Compare them to identify binding ceilings rather than assuming
all four inputs must equal their limits. There are no tax payments.

```{r}
#| label: hh-11-1
# A common percentage r creates four soil-specific ceilings.
both_caps_gap <- function(r) {
  x <- choose_inputs(ceilings = (1-r) * baseline_inputs)
  mean_leaching(x) - leaching_target
}
reduction_fraction <- uniroot(both_caps_gap, interval = c(0, 0.3), tol = 1e-9)$root
reduction_fraction
both_ceilings <- (1-reduction_fraction) * baseline_inputs
both_ceilings
both_caps_inputs <- choose_inputs(ceilings = both_ceilings)
farm_results(both_caps_inputs)
both_caps_result <- policy_result("Water and nitrogen caps (% of baseline)", both_caps_inputs)
```

## 12. Compare policies

`rbind()` stacks the saved one-row data frames into a single table. The
baseline plus seven policies produces eight rows. `digits = 6` controls
printed precision, while `row.names = FALSE` hides row labels.

| Column | How to interpret it |
|---|---|
| `efficiency_cost` | Lost pre-tax quasi-rent relative to baseline, dollars/ha |
| `leaching` | Average leaching, kg/ha |
| `tax_payments` | Area-weighted payments to the regulator, dollars/ha |
| `farmer_net_rent` | Quasi-rent after tax payments, dollars/ha |

First compare each regulated row's leaching with `leaching_target`, allowing
for numerical tolerance. Then compare efficiency costs at that common
outcome. Finally, compare farmer net rents to see the additional effect of
tax payments. The efficient benchmark should provide a lower bound on
production costs for the other feasible policies in this model; a lower
reported cost elsewhere is a reason to inspect feasibility and convergence.

The last two calls return to soil-level outcomes. An aggregate table can
hide which soil adjusts and which farmers bear the production losses.

```{r}
#| label: hh-12-1
policy_comparison <- rbind(baseline_result, efficient_result, uniform_result,
                           both_caps_result, nitrogen_tax_result, water_tax_result,
                           nitrogen_cap_result, water_cap_result)
print(policy_comparison, digits = 6, row.names = FALSE)

# Inspect any soil-level allocation using farm_results(), for example:
farm_results(water_tax_inputs)
farm_results(water_cap_inputs)

# This walkthrough intentionally omits the original program's checks,
# full derivative machinery, multi-start searches, and automatic root bracketing.
# Keep model.R and validate_model.R as the reference for research/extensions.
```

### Questions to answer from your run

1. Why do the two soils choose different inputs at the same market prices?
2. How do the differentiated nitrogen and water taxes vary across soils,
   and how does marginal leaching explain those differences?
3. How much additional efficiency cost comes from restricting the regulator
   to uniform rates or to a single taxed input?
4. Which ceilings bind? How do adjustments in the unrestricted input affect
   the pollution outcome?
5. Which policy has the lowest efficiency cost, and which leaves farmers
   with the highest net rent? Explain any difference using the tax payments.

### If you change the model

Run again from Section 1 so all functions, the baseline, the target, and
policy results reflect the same assumptions. Recheck root-search brackets,
solver messages, and target attainment. SLSQP is a local numerical method;
its output alone does not prove global optimality in an altered model.

`HH_model.R` deliberately keeps the classroom steps visible and omits the
larger implementation's automated verification and multiple-start searches.
Use `model.R` and `validate_model.R` as the reference for research extensions.
The separate [model lab](Helfand_Model_Lab.qmd) provides the assignment.

## Reference

Helfand, Gloria E., and Brett W. House. 1995. "Regulating Nonpoint Source
Pollution Under Heterogeneous Conditions." *American Journal of Agricultural
Economics* 77(4):1024–1032. <https://doi.org/10.2307/1243825>.
